### 2016-06-14 PaperOnSmoothing/draft-31052016.tex
Add files via upload
Initial tex and bib files for the paper.

\documentclass[10pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fullpage}
\begin{document}

\title{Crowd density estimation from Wi-Fi positioning data in the Amsterdam Arena}
\maketitle

\begin{abstract}
We present the Amsterdam Arena project, which involves observing and managing crowd behaviour, using the Amsterdam Arena stadium as a living laboratory. The main scientific question we explore is how to detect anomalous behaviour in large crowds in real time. The main purpose is to be able to predict a possible crowd disaster and identify means to prevent it from happening. 
Human crowds are complex systems, and predicting or controlling their behaviour is challenging. Our approach involves three phases, firstly data collection from Wi-Fi and Bluetooth sensors in the stadium, secondly data analytics, and finally we aim to use simulation to make forecasts about the crowd dynamics. 
Here we present initial results on the data analytics and show how we can extract density maps from Wi-Fi positioning data. The technology we deploy is based on the Wi-Fi signals from smart phones. We use the existing network of Wi-Fi access points in the stadium, and capture probe signals from smart phones, which are processed and anonymised in alignment with privacy concerns. The positions of smart phones are reconstructed using received signal strengths and methods similar to trilateration. The data provides us with real-time information on spatial distributions of crowd density, which is an essential indicator of the criticality of crowd conditions. To visualise the spatiotemporal behaviour of crowd density in real-time we dynamically generate heat maps along a moving time interval. The heat maps are generated through the statistical modelling of the positioning data. The generation of dynamic heat maps allows us to detect and locate hot-spots of density where crowd conditions reach critical values that could possibly lead to a disaster.
\end{abstract}

\section{Introduction}

We present the Amsterdam Arena project, which involves observing and managing crowd behavior, using the Amsterdam Arena stadium as a living laboratory. The main scientific question we explore is how to detect anomalous behavior in large crowds in real time.

\begin{itemize}

\item Background\\

Human crowds are complex systems, and predicting or controlling their behavior is challenging. Our approach involves three phases, firstly data collection from Wi-Fi and Bluetooth sensors in the stadium, secondly data analytics, and finally we aim to use simulation to make forecasts about the crowd dynamics.

In this paper we present initial results on the data analytics and show how we can extract density maps from Wi-Fi positioning data.

\item Aims\\

The main purpose is to be able to predict a possible crowd disaster and identify means to prevent it from happening.

\item Related work\\

Measuring crowd density by tracking the location of smart phones has been done in:

(Wirz \textit{et al.} 2012) \cite{wirz:1} and (Wirz \textit{et al.} 2013) \cite{wirz:2} use GPS data of people that use an App.\\
(Weppner \textit{et al.} 2014) \cite{weppner:1} use Bluetooth, however without positioning / tracking.\\
(Schauer \textit{et al.} 2014) \cite{schauer:1} use Wi-Fi, without positioning / tracking.

\end{itemize}

\section{Data}

\begin{itemize}
\item Wi-Fi positioning
\end{itemize}

The technology we deploy is based on the Wi-Fi signals from smart phones. We use the existing network of Wi-Fi access points in the stadium, and capture probe signals from smart phones, which are processed and anonymised in alignment with privacy concerns. The positions of smart phones are reconstructed using received signal strengths and methods similar to trilateration.

\begin{itemize}
\item How are the uncertainties (errors) generated and what do they mean?
\end{itemize}

\section{Methods}

The data provides us with real-time information on spatial distributions of crowd density, which is an essential indicator of the criticality of crowd conditions.

\subsection{Density estimation}

To visualise the spatiotemporal behaviour of crowd density in real-time we dynamically generate heat maps along a moving time interval. The heat maps are generated through the statistical modeling of the positioning data. The generation of dynamic heat maps allows us to detect and locate hot-spots of density where crowd conditions reach critical values that could possibly lead to a disaster.

\begin{itemize}
\item The method we use is similar to Kernel Density Estimation, but what we are doing is not actual smoothing (?).
\end{itemize}

To construct two-dimensional probability density functions we apply non-parametric density estimation.
To make density histograms we consider a two-dimensional binned region of space, and count the number of positions that fall into each bin. For each bin location $X$, in the time interval $t$, we apply the kernel density estimation (KDE) method \cite{scott}\cite{silverman} given by

\begin{equation}
\hat{d}(X,t)=\frac{1}{N }\sum_{i=1}^{N} K_{h_{x}} (x-x_{i}) K_{h_{y}} (y-y_{i})
\label{kde}
\end{equation}

where $K_{h_{u}}$ is the kernel function $K_{h}(u)=K(u/h)/h$, $h$ is the smoothing parameter, or bandwidth, and $N$ is the number of data points (positions) in the time interval $t$.

The bandwith parameter $h$ is crucial for the accuracy of the density estimate. 
Several theoretical approaches are possible.
Here we base our choice of kernel bandwidth values $h$ on the error values $(\sigma_{x},\sigma_{y})$ provided by the positioning methods (see Section ).

\subsection{Real-time data analysis}

To process positioning data in real-time during events, we consider data within a specific moving time window.
To take in account that visitors are moving, we subdivide the time window in multiple sub-windows and attach more weight to measurements in later sub-windows. The weighting scheme is given by

\begin{equation}
\hat{d}(X,t)=\frac{w_{1}\hat{f}(X,t_{1})+w_{2}\hat{f}(X,t_{2})+...+w_{m}\hat{f}(X,t_{m})}{w_{1}N_{1}+w_{2}N_{2}+...+w_{m}N_{m}}
\end{equation}

\noindent where $\hat{f}(X,t)$ is the non-normalized sum of smoothed counts ($N\times\hat{d}(X,t)$) in Equation \ref{kde}, $w_{k}$ is the weight value given to sub-window $k$, and $m$ is the number of sub-windows.

\subsection{Systematic evaluation}

\begin{itemize}
\item Simulation?\\
Do we include simulation of the positioning process, i.e. the toy Monte Carlo simulation, or do we generate fitted positions with error values, and only test the density estimation methods?
\end{itemize}

\section{Results}

\section{Discussion}

\begin{itemize}
\item Future work
\item Conclusion
\end{itemize}

\bibliography{references}
\bibliographystyle{plain}

\end{document}

### 2016-06-21 PaperOnSmoothing/draft-31052016.tex

@@
 
+\end{itemize}
+
 Measuring crowd density by tracking the location of smart phones has been done in:
@@
 (Weppner \textit{et al.} 2014) \cite{weppner:1} use Bluetooth, however without positioning / tracking.\\
-(Schauer \textit{et al.} 2014) \cite{schauer:1} use Wi-Fi, without positioning / tracking.
+(Schauer \textit{et al.} 2014) \cite{schauer:1} use Wi-Fi, without positioning / tracking.\\
 
-\end{itemize}
+Wirz \textit{et al.} (2012) \cite{wirz:1} infer crowd density from GPS location traces of people who use an App during the 2011 Lord Mayor Show in London.
+They visualize the data in real-time, and generate heat maps (dynamically) using the kernel density estimation (KDE) method.
+In (Wirz \textit{et al.} 2013) \cite{wirz:2} the methodology is presented more elaborately, using a real-world data set collected during the same festival.
+The relation between the number of App users and the crowd density is calibrated (using linear regression), which forms the basis of their participatory sensing method.
+The accuracy of the smoothing method is assessed using ground truth crowd density information obtained from video footage recorded by CCTV cameras. 
+The influence of the kernel radius on the correlation between the actual density and the density of App is analysed.
+
+The Gaussian weight function is introduced in Helbing \textit{et al.} (2007) \cite{helbing:1} to estimate local densities in an area captured by video cameras during the Hajj in Mina/Makkah in 1426H on January 12, 2006.
 

### 2016-06-21 PaperOnSmoothing/draft-31052016.tex
introduction started

@@
 
-We present the Amsterdam Arena project, which involves observing and managing crowd behavior, using the Amsterdam Arena stadium as a living laboratory.
-The main scientific question we explore is how to detect anomalous behavior in large crowds in real time.
+***********************************From Sonja. nb: the text will be polished in the final phases, first I care only about putting all ideas on paper***************
 
+Crowd disasters have taken many human lives.
+The Love Parade disaster (Duisburg, 2010), the Ellis Park Stadium disaster (Johannesburg, 2001), the PhilSports Stadium stampede (Manila, 2006) are just a few recent examples.
+One of the reasons that disasters happen is a "lack of overview of everybody" [cite paper on love parade EPJ DS], that is, a lack of macroscopic overview of the crowd.
+Critical crowd density (cite paper 8/m2, find it in e-mails)) is a contributing factor to crowd disasters; yet, it is still challenging to determine timely when it occurs, so that a disaster can be prevented before it happens.
+
+In our work we are interested in estimating crowd density during concerts in indoor spaces.
+A lot of research on estimating crowd density has been done using video processing from security cameras (cite papers (1), (2), (3) from proposal?).
+However, this approach does not apply in our case, because 1)  it is difficult to obtaine macroscopic overview ; 2) the lighting conditions during concert hours are not sufficient for video-based crowd analysis; and 3) the error of video-based density estimation increases with the increase of the actual crowd density.
+
+In our approach, we combat the three above mentioned issues with video analysis by exploiting the ubiquity of smart phones, as in ((4) (5), (6) from proposal?).
+Our living laboratory is the Amsterdam Arena(citatioon).
+First, we are estimating in real-time anonymously the positions of the visitors via the signal strengths of their phones to the  Wi-Fi routers, as presented in (cite Jan's slides), via a technique similar to trilateration.
+Second, we apply the law of the large numbers to obtain an estimation of the crowd density, also in real-time.
+With the first step, we are dealing with issues 1) and 2) mentioned above.
+With the second step, we are handling issue 3) -  that is, with our method the precision of estimation actually increases as the crowd density increases.
+
+n.b.
+in related work section put all that use wi-fi tracking, for those that use GPS tracking only mention that this does not work in indoor spaces and that it requires participation.
+Not necessary to put work on video tracking in related work - it is already mentioned in intro.
+Mention also bluetooth tracking, but bluetooth is not so ubiquitous and our principles apply also if bluetooth was used (right?)
+
+Finally, none of the mentioned work uses our approach of modeling the position of an individual as a probability distribution; our method is designed to attack the problem of having a dense crowd; we are not interested in precise estimations for freely moving crowds; rather, we are interested in obtaining precise estimation when the  concert crowds is dense and static due to this.
+blah blah (sonja: explain better)
+
+************************************end from Sonja
 \begin{itemize}

### 2016-06-22 PaperOnSmoothing/draft-31052016.tex
introduction

@@
 One of the reasons that disasters happen is a "lack of overview of everybody" [cite paper on love parade EPJ DS], that is, a lack of macroscopic overview of the crowd.
-Critical crowd density (cite paper 8/m2, find it in e-mails)) is a contributing factor to crowd disasters; yet, it is still challenging to determine timely when it occurs, so that a disaster can be prevented before it happens.
+Critical crowd density (cite paper 8/m2, find it in e-mails)) is a contributing factor to crowd disasters; yet, it is still challenging to determine timely when it occurs, so that a disaster can be prevented before it happens, by e.g.
+navigating the rest of the crowd away from the congestion.
 
@@
 A lot of research on estimating crowd density has been done using video processing from security cameras (cite papers (1), (2), (3) from proposal?).
-However, this approach does not apply in our case, because 1)  it is difficult to obtaine macroscopic overview ; 2) the lighting conditions during concert hours are not sufficient for video-based crowd analysis; and 3) the error of video-based density estimation increases with the increase of the actual crowd density.
+However, this approach does not apply in our case, because 1)  it is difficult to obtain macroscopic overview ; 2) the lighting conditions during concert hours are not sufficient for video-based crowd analysis; and 3) the error of video-based density estimation increases with the increase of the actual crowd density.
 
-In our approach, we combat the three above mentioned issues with video analysis by exploiting the ubiquity of smart phones, as in ((4) (5), (6) from proposal?).
-Our living laboratory is the Amsterdam Arena(citatioon).
-First, we are estimating in real-time anonymously the positions of the visitors via the signal strengths of their phones to the  Wi-Fi routers, as presented in (cite Jan's slides), via a technique similar to trilateration.
-Second, we apply the law of the large numbers to obtain an estimation of the crowd density, also in real-time.
+In our approach, we combat the three above mentioned issues by exploiting the ubiquity of smart phones, as in ((4) (5), (6) from proposal?).
+Our living laboratory is the Amsterdam Arena(citation).
+First, we are estimating in real-time anonymously the positions of the visitors via the signal strengths of their phones to the  Wi-Fi routers in the venue.
+This technique has been presented in (cite Jan's slides), and is similar to trilateration.
+However, due to signal interference, irregular internet packet rates and systematic errors, this methodology is imprecise as well.
+Thus, second, we apply the law of the large numbers to obtain an estimation of the crowd density, also in real-time.
 With the first step, we are dealing with issues 1) and 2) mentioned above.
-With the second step, we are handling issue 3) -  that is, with our method the precision of estimation actually increases as the crowd density increases.
+With the second step, which is the main contribution of this paper, we are handling issue 3) -  that is, with our method the precision of estimation actually increases as the crowd density increases.
 

### 2016-06-22 PaperOnSmoothing/draft-31052016.tex
paper re-organized, related work moved before conclusion, some parts of text commented

@@
 \usepackage{fullpage}
+\usepackage{verbatim}
 \begin{document}
@@
 \begin{abstract}
+\begin{comment}
 We present the Amsterdam Arena project, which involves observing and managing crowd behaviour, using the Amsterdam Arena stadium as a living laboratory.
@@
 \end{abstract}
-
+\end{comment}
 \section{Introduction}
@@
 ************************************end from Sonja
+\begin{comment}
 \begin{itemize}
@@
 
-\item Related work\\
+\end{itemize}
+\end{comment}
 
-\end{itemize}
+\section{Positioning of visitors using Wi-Fi sensors and smart phones}
 
-Measuring crowd density by tracking the location of smart phones has been done in:
-
-(Wirz \textit{et al.} 2012) \cite{wirz:1} and (Wirz \textit{et al.} 2013) \cite{wirz:2} use GPS data of people that use an App.\\
-(Weppner \textit{et al.} 2014) \cite{weppner:1} use Bluetooth, however without positioning / tracking.\\
-(Schauer \textit{et al.} 2014) \cite{schauer:1} use Wi-Fi, without positioning / tracking.\\
-
-Wirz \textit{et al.} (2012) \cite{wirz:1} infer crowd density from GPS location traces of people who use an App during the 2011 Lord Mayor Show in London.
-They visualize the data in real-time, and generate heat maps (dynamically) using the kernel density estimation (KDE) method.
-In (Wirz \textit{et al.} 2013) \cite{wirz:2} the methodology is presented more elaborately, using a real-world data set collected during the same festival.
-The relation between the number of App users and the crowd density is calibrated (using linear regression), which forms the basis of their participatory sensing method.
-The accuracy of the smoothing method is assessed using ground truth crowd density information obtained from video footage recorded by CCTV cameras. 
-The influence of the kernel radius on the correlation between the actual density and the density of App is analysed.
-
-The Gaussian weight function is introduced in Helbing \textit{et al.} (2007) \cite{helbing:1} to estimate local densities in an area captured by video cameras during the Hajj in Mina/Makkah in 1426H on January 12, 2006.
-
-\section{Data}
-
-\begin{itemize}
-\item Wi-Fi positioning
-\end{itemize}
-
+\begin{comment}
 The technology we deploy is based on the Wi-Fi signals from smart phones.
@@
 \end{itemize}
+\end{comment}
+\section{From positioning to estimation of density}
 
-\section{Methods}
+\subsection{Issues with positioning (title to reconsider)}
 
-The data provides us with real-time information on spatial distributions of crowd density, which is an essential indicator of the criticality of crowd conditions.
+\subsection{Addressing the issues: real-time crowd density estimation}
 
-\subsection{Density estimation}
-
+\begin{comment}
 To visualise the spatiotemporal behaviour of crowd density in real-time we dynamically generate heat maps along a moving time interval.
@@
 The generation of dynamic heat maps allows us to detect and locate hot-spots of density where crowd conditions reach critical values that could possibly lead to a disaster.
+\end{comment}
 
@@
 
-\subsection{Systematic evaluation}
+\subsection{Validation}
 
@@
 
-\section{Results}
+\section{Related work}
+Measuring crowd density by tracking the location of smart phones has been done in:
 
-\section{Discussion}
+(Wirz \textit{et al.} 2012) \cite{wirz:1} and (Wirz \textit{et al.} 2013) \cite{wirz:2} use GPS data of people that use an App.\\
+(Weppner \textit{et al.} 2014) \cite{weppner:1} use Bluetooth, however without positioning / tracking.\\
+(Schauer \textit{et al.} 2014) \cite{schauer:1} use Wi-Fi, without positioning / tracking.\\
 
-\begin{itemize}
-\item Future work
-\item Conclusion
-\end{itemize}
+Wirz \textit{et al.} (2012) \cite{wirz:1} infer crowd density from GPS location traces of people who use an App during the 2011 Lord Mayor Show in London.
+They visualize the data in real-time, and generate heat maps (dynamically) using the kernel density estimation (KDE) method.
+In (Wirz \textit{et al.} 2013) \cite{wirz:2} the methodology is presented more elaborately, using a real-world data set collected during the same festival.
+The relation between the number of App users and the crowd density is calibrated (using linear regression), which forms the basis of their participatory sensing method.
+The accuracy of the smoothing method is assessed using ground truth crowd density information obtained from video footage recorded by CCTV cameras. 
+The influence of the kernel radius on the correlation between the actual density and the density of App is analysed.
+
+The Gaussian weight function is introduced in Helbing \textit{et al.} (2007) \cite{helbing:1} to estimate local densities in an area captured by video cameras during the Hajj in Mina/Makkah in 1426H on January 12, 2006.
+
+\section{Conclusion}
 

### 2016-06-22 PaperOnSmoothing/draft-31052016.tex
Added some more to the Related Work section.

@@
 The generation of dynamic heat maps allows us to detect and locate hot-spots of density where crowd conditions reach critical values that could possibly lead to a disaster.
+\end{comment}
 \end{abstract}
-\end{comment}
+
 \section{Introduction}
@@
 \section{Related work}
-Measuring crowd density by tracking the location of smart phones has been done in:
-
-(Wirz \textit{et al.} 2012) \cite{wirz:1} and (Wirz \textit{et al.} 2013) \cite{wirz:2} use GPS data of people that use an App.\\
-(Weppner \textit{et al.} 2014) \cite{weppner:1} use Bluetooth, however without positioning / tracking.\\
-(Schauer \textit{et al.} 2014) \cite{schauer:1} use Wi-Fi, without positioning / tracking.\\
 
@@
 
-The Gaussian weight function is introduced in Helbing \textit{et al.} (2007) \cite{helbing:1} to estimate local densities in an area captured by video cameras during the Hajj in Mina/Makkah in 1426H on January 12, 2006.
+The Gaussian weight function used in (Wirz \textit{et al.} 2012; 2013) \cite{wirz:1}\cite{wirz:2} is introduced in Helbing \textit{et al.} (2007) \cite{helbing:1} to estimate local densities in an area captured by video cameras during the Hajj in Mina/Makkah in 1426H on January 12, 2006.\\
+
+A number of studies use Bluetooth to estimate crowd densities at a wide range of places and events.
+In these studies the position of a mobile phone is approximated to the location of the sensor by which it is detected.
+
+Schauer \textit{et al.} (2014) \cite{schauer:1} count unique MAC addresses detected by two sensors (nodes) at both sides (public and security) of a security check inside a major German airport, to estimate pedestrian densities and pedestrian flow.
+They consider time information, to determine the direction of a person's movement, and at least one RSSI value, to reduce the number of false positives in case devices are captured by both sensors.
+They compare Bluetooth and Wi-Fi based methods, and compare their methods to a known ground truth provided by the number of security checks.
+
+Versichele \textit{et al.} (2012) \cite{versichele:1} use Bluetooth scanners at strategic locations during the 10-day Ghent Festivities, to analyze spatio-temporal dynamics of pedestrians.
+
+Yoshimura \textit{et al.} (2016) \cite{yoshimura:1} use Bluetooth detection to analyze visitors' behavior at the Louvre museum in Paris.
+
+Delafontaine \textit{et al.} (2012) \cite{delafontaine:1} use a similar approach (of Bluetooth tracking) and apply (genetic) sequence alignment methods to analyze the resulting data which consists of different sequences of sensors (nodes) for detected mobile devices.
+
+Weppner and Lukowicz (2013) \cite{weppner:1} estimate crowd densities during a soccer European championship public viewing event and during the Oktoberfest, by distributing volunteers in the crowd, who are carrying smart phones scanning for Bluetooth devices.
+They then use statistics to combine the different measurements in space and time.
+
+In Weppner \textit{et al.} (2014) \cite{weppner:2} similar methods are applied to a city-wide festival in Zurich 2013, now supported by GPS data.
 

### 2016-06-24 PaperOnSmoothing/draft-31052016.tex
introduction updated. issues with positioning section started

@@
 \usepackage{verbatim}
+\usepackage{graphicx}
 \begin{document}
@@
 Our living laboratory is the Amsterdam Arena(citation).
-First, we are estimating in real-time anonymously the positions of the visitors via the signal strengths of their phones to the  Wi-Fi routers in the venue.
-This technique has been presented in (cite Jan's slides), and is similar to trilateration.
-However, due to signal interference, irregular internet packet rates and systematic errors, this methodology is imprecise as well.
-Thus, second, we apply the law of the large numbers to obtain an estimation of the crowd density, also in real-time.
-With the first step, we are dealing with issues 1) and 2) mentioned above.
-With the second step, which is the main contribution of this paper, we are handling issue 3) -  that is, with our method the precision of estimation actually increases as the crowd density increases.
+Or methodology consists of three steps.
+First, we are using the Wi-Fi routers in the stadium to gather the strengths of the wireless antena signals from individuals smart phones.
+Second, we are estimating anonymously the positions of the visitors from the signal strengths in real time.
+This technique has been presented in (cite Jan's slides), and is similar to trilateration, that is, the way the Global Positioning System (GPS) is implemented.
+However, due to signal interference and blockage in dense crowds, irregular internet packet rates and systematic errors leading to multiple local optima, the positioning itself does not suffice to estimate the crowd density.
+Thus, in our third step, we aggregate the results from the positioning and apply the law of the large numbers to obtain an estimation of the crowd density, also in real-time.
+With the first two steps, we are dealing with issues 1) and 2) mentioned above.
+With the third step, which is the main contribution of this paper, we are handling issue 3) -  that is, with our method the precision of estimation actually increases as the crowd density increases.
+
+The rest of the paper is organized as follows.....*******
 
@@
 
+This section will be written later, maybe with input from Jan.
+
 \begin{comment}
@@
 
+Sonja:Mention somewhere that in the documentary of the LP disaster people were raising their phones up to look for a way out. (which is good , less blockage of signals)
 \subsection{Issues with positioning (title to reconsider)}
+
+Under ideal circumstances, the positioning itself would suffice to estimate the crowd density per region of interest:  every few seconds we would only need to count the number of estimates that fall in the region of interest.
+However, we have observed several issues, which are not related to the mathematical methodology behind the positioning step, that prevent us to apply direct counting.
+\begin{enumerate}
+	\item
+	Due to signal interference, when the optimization of the estimation of location of a certain MAC address is performed, there can be multiple local optima that are equally good candidates.
+For example, consider Fig.
+\ref{fig:trilateration_problem}.
+In the center of every ring there is a Wi-Fi router, that has estimated signal strength to a certain MAC device with a certain error range.
+The error range is represented by the thickness of the ring.
+Then, there are two possible regions which are equally good candidates for the positioning of the MAC device, and those regions are the two darkest regions where all three rings overlap.
+Note that the problem can arise regardless of the thickness of the rings or the number of Wi-Fi routers.
+	\begin{figure}[h!]
+		\centering
+		\includegraphics[width=40mm]{trilateration_problem.jpg}
+		\caption{Trilateration can lead to multiple local optima}
+		\label{fig:trilateration_problem}
+	\end{figure}
+	In fact, we have taken a static MAC device and plotted the estimations of its coordinates through time, where we have plotted only the estimations with a relatively small error  (Fig.
+\ref{fig:bimodal})
+	\begin{figure}[h!]
+		\centering
+		\includegraphics[width=100mm]{bimodal.jpg}
+		\caption{Trilateration can lead to multiple local optima (cut the lower part from the picture)}
+		\label{fig:bimodal}
+	\end{figure} 
+	The resulting plot (for the x-coordinate) reveals a bimodal distribution of the estimation. This phenomenon can be explained by the multiple local optima situation.
+
+\item Volatility of packet rates.
+When a MAC device is connected to the WI-Fi internet, it sends packets with a relatively stable rate.
+However, during concert hours, very few devices are actually connected to the internet, and most of our estimations of positions come from signals that the device sends in a "probing" mode.
+In this case the packet rate is quite volatile, ranging from a few seconds to a few minutes (some citation?).
+\end{enumerate}
 

### 2016-06-24 PaperOnSmoothing/draft-31052016.tex
issues section updated

@@
 
-n.b.
+*************n.b.
 in related work section put all that use wi-fi tracking, for those that use GPS tracking only mention that this does not work in indoor spaces and that it requires participation.
@@
 Mention also bluetooth tracking, but bluetooth is not so ubiquitous and our principles apply also if bluetooth was used (right?)
-
 Finally, none of the mentioned work uses our approach of modeling the position of an individual as a probability distribution; our method is designed to attack the problem of having a dense crowd; we are not interested in precise estimations for freely moving crowds; rather, we are interested in obtaining precise estimation when the  concert crowds is dense and static due to this.
-blah blah (sonja: explain better)
+blah blah (sonja: explain better)********
 
@@
 	\item
+	{\it Multiple local optima}.
 Due to signal interference, when the optimization of the estimation of location of a certain MAC address is performed, there can be multiple local optima that are equally good candidates.
 For example, consider Fig.
-\ref{fig:trilateration_problem}.
+\ref{fig:trilateration_problem} \footnote{Picture courtesy of http://math.stackexchange.com/questions/42537/trilateration-with-bounds}.
 In the center of every ring there is a Wi-Fi router, that has estimated signal strength to a certain MAC device with a certain error range.
@@
 	\end{figure}
-	In fact, we have taken a static MAC device and plotted the estimations of its coordinates through time, where we have plotted only the estimations with a relatively small error  (Fig.
+	In fact, let us consider a static (24 hours persistent) MAC device.
+We have  plotted the estimations of its coordinates through time, where we have plotted only the estimations with a relatively small error  (Fig.
 \ref{fig:bimodal})
@@
 	\end{figure} 
-	The resulting plot (for the x-coordinate) reveals a bimodal distribution of the estimation. This phenomenon can be explained by the multiple local optima situation.
+	The resulting plot (for the x-coordinate) reveals a bimodal distribution of the estimation.
+We have observed such a distribution in a majority of the 20 randomly sampled MAC addresses.
+This phenomenon can be explained by the multiple local optima situation, because we did not observe bi-modal distributions in the signal strengths.
 
-\item Volatility of packet rates.
+\item {\it Volatility of packet rates}.
 When a MAC device is connected to the WI-Fi internet, it sends packets with a relatively stable rate.
@@
 In this case the packet rate is quite volatile, ranging from a few seconds to a few minutes (some citation?).
+This means that when we make a snapshot of the MAC addresses visible in a certain moment, we are only detecting a small fraction of the MAC devices.
+
+\item {\it MAC randomization}.
+Due to ever increasing privacy concerns and possibly other business reasons, starting from 2014, the Apple I-phones have introduced randomization of the MAC address when the device is in a probing mode (not connected to internet) (citation?).
+This means that when the device switches from a probing to an online mode or vice versa, the MAC address suddenly changes and thus the device cannot be followed anymore.
+(n.b.
+here or later: there is a tag that reveals whether the address is randomized or not)
 \end{enumerate}

### 2016-06-24 PaperOnSmoothing/draft-31052016.tex
addressing the issues section: started writing

@@
 This phenomenon can be explained by the multiple local optima situation, because we did not observe bi-modal distributions in the signal strengths.
+(***** Sonja: better to first write the observation of bimodality and then the multiple local optima possible explanation***)
 
@@
 \subsection{Addressing the issues: real-time crowd density estimation}
+
+*************from Sonja
+
+In order to deal with issue 1 from above, let us start with the following observation: we are not interested in the individual locations of the MAC devices, but rather in the density of the crowd; we would like especially to have more precise estimation when the density reaches a certain threshold, so that the crowd that is not in the dense region can be alarmed promptly to move away from the dense region.
+So, we are interested in estimating density accurately when the crowd is not in a state of movement (because if people are able to freely move, they are not in a danger).
+This situation, dense crowd, means that there is more blockage of signals from bodies, less precise estimation of distance from Wi-Fi based on signal strength, which means that the rings in Fig.\ref{fig:trilateration_problem} would be even thicker.
+This exaggerates the effect of "bimodal" or even multi-modal distribution of position estimates.
+Thus, in a dense crowd, all we can derive from the positioning method for a MAC address is a probability distribution of its possible locations.
+For example, in the scenario in Fig.
+\ref{fig:trilateration_problem}, we can say that the MAC device is with a probability of 0.5 in the left grey region (region where all three rings overlap) and with a probability of 0.5 in the right grey region.
+While this does not provide us with useful information for the location of the MAC address, if we apply the same reasoning for all MAC addresses, and, for a region R we add together the probabilities for all MAC devices that they are in that region, we end up with an estimation of the number of MAC devices in that region.
+If we assume that the probability distributions of the locations of the MAC addresses are independent, we can directly apply the law of the large numbers and conclude that for a dense crowd the error of the estimation of the density per region will vanish.
+However, we cannot assume the mentioned independence, because people tend to go in groups to concerts.
+In this case, the variance of the estimation is equal to the average correlation between the distributions in the limiting case.
+Nevertheless, we can safely assume that for a regular crowd at a concert the average correlation tends to zero as the number of people increases; thus, the error of the estimation of the crowd density will in anyway vanish as the number of people tends to infinity.
+Formally, let R be an arbitrary region.
+(now I have to write some formulas, will continue on Monday)....
+
+***************end from Sonja
+
+***start from Philip
 

### 2016-06-27 PaperOnSmoothing/draft-31052016.tex
Started writing on Methods section.

@@
 To construct two-dimensional probability density functions we apply non-parametric density estimation.
-To make density histograms we consider a two-dimensional binned region of space, and count the number of positions that fall into each bin.
 For each bin location $X$, in the time interval $t$, we apply the kernel density estimation (KDE) method \cite{scott}\cite{silverman} given by
 
+The data provides us with a series of estimated positions, together with corresponding values for the uncertainty (see Section on Positioning).
+From the estimated positions we wish to generate density histograms.
+
+To make density estimates we consider a two-dimensional binned region of space, and count the number of positions that fall into each bin.
+We wish to visualize / preserve the uncertainty for each estimated position.
+Therefore we transform each bin count into a bivariate Gaussian, using the separate error values ($\sigma_{x}$ and $\sigma_{y}$) for each dimension / direction.
+For each MAC we construct a two-dimensional density function that is the sum of these smooth positions.
+Because the number of measurements are different for each detected MAC address (in the interval) we normalize each histogram separately.
+
+We sum all the density histograms to create the total probability density histogram, which should ideally integrate to the number of MAC's detected (in the interval).
+
+We smooth each position by evaluating the following function at each bin in the histogram,
+
 \begin{equation}
-\hat{d}(X,t)=\frac{1}{N }\sum_{i=1}^{N} K_{h_{x}} (x-x_{i}) K_{h_{y}} (y-y_{i})
+\hat{d}(x,y)=K((x-x_{i}),\sigma_{x}) K((y-y_{i}),\sigma_{y})
 \label{kde}
 \end{equation}
-
-where $K_{h_{u}}$ is the kernel function $K_{h}(u)=K(u/h)/h$, $h$ is the smoothing parameter, or bandwidth, and $N$ is the number of data points (positions) in the time interval $t$.
-
-The bandwith parameter $h$ is crucial for the accuracy of the density estimate. 
-Several theoretical approaches are possible.
-Here we base our choice of kernel bandwidth values $h$ on the error values $(\sigma_{x},\sigma_{y})$ provided by the positioning methods (see Section ).
+where the location $(x,y)$ is expressed in bin unit, and where we use the kernel function
+\begin{equation}
+K(u,\sigma)=\frac{1}{\sigma\sqrt{2\pi}}\exp(\frac{u^2}{2\sigma^2})
+\end{equation}
+Because the distances are expressed in bin units we scale the sigma's  ($\sigma_{x}$ and $\sigma_{y}$) by dividing the error values $\epsilon$ by the bin size $h$ 
+\begin{equation}
+\sigma=\frac{\epsilon}{h}
+\end{equation}
+This implies that the bin size becomes the unit for our density estimates, and we generate probabilities for number of people per bin, which has the area $h\times h$.
 

### 2016-06-28 PaperOnSmoothing/draft-31052016.tex
methods section updated

@@
 	\item
-	{\it Multiple local optima}.
-Due to signal interference, when the optimization of the estimation of location of a certain MAC address is performed, there can be multiple local optima that are equally good candidates.
+	{\it Bimodal distributions of coordinates estimations}.
+We have sampled randomly 20 MAC addresses and plotted the estimations of their coordinates through time, where we have plotted only the estimations with a relatively small uncertainty.
+We observed persistent bimodal distribution of the estimations, an example of which can be seen in Fig.
+\ref{fig:bimodal}.
+		\begin{figure}[h!]
+			\centering
+			\includegraphics[width=100mm]{bimodal.jpg}
+			\caption{Trilateration can lead to multiple local optima (cut the lower part from the picture)}
+			\label{fig:bimodal}
+		\end{figure} 
+	This figure shows the estimated x-coordinates through time of a static, 24 hours persistent MAC address.
+Because we did not observe  bi-modal distributions in the signal strengths, the bi-modal distribution of the positions estimates can be explained by the multiple local optima situation.
+Namely, due to signal interference, when the optimization of the estimation of location of a certain MAC address is performed, there can be multiple local optima that are equally good candidates.
 For example, consider Fig.
@@
 	\end{figure}
-	In fact, let us consider a static (24 hours persistent) MAC device.
-We have  plotted the estimations of its coordinates through time, where we have plotted only the estimations with a relatively small error  (Fig.
-\ref{fig:bimodal})
-	\begin{figure}[h!]
-		\centering
-		\includegraphics[width=100mm]{bimodal.jpg}
-		\caption{Trilateration can lead to multiple local optima (cut the lower part from the picture)}
-		\label{fig:bimodal}
-	\end{figure} 
-	The resulting plot (for the x-coordinate) reveals a bimodal distribution of the estimation.
-We have observed such a distribution in a majority of the 20 randomly sampled MAC addresses.
-This phenomenon can be explained by the multiple local optima situation, because we did not observe bi-modal distributions in the signal strengths.
-(***** Sonja: better to first write the observation of bimodality and then the multiple local optima possible explanation***)
 
@@
 \begin{comment}
+
 To visualise the spatiotemporal behaviour of crowd density in real-time we dynamically generate heat maps along a moving time interval.
@@
 The generation of dynamic heat maps allows us to detect and locate hot-spots of density where crowd conditions reach critical values that could possibly lead to a disaster.
-\end{comment}
-
-\begin{itemize}
-\item The method we use is similar to Kernel Density Estimation, but what we are doing is not actual smoothing (?).
-\end{itemize}
 
@@
 For each bin location $X$, in the time interval $t$, we apply the kernel density estimation (KDE) method \cite{scott}\cite{silverman} given by
+
+\end{comment}
+
+{\it below: emphasize that we measure in a time window.
+The final result of stacking is not a probability density distribution but a crowd density distribution.
+We don't generate probabilities for number of people per bin, but we generate an estimation of number of people per bin.
+}
 
@@
 
+\begin{comment}
 \subsection{Real-time data analysis}
@@
 \noindent where $\hat{f}(X,t)$ is the non-normalized sum of smoothed counts ($N\times\hat{d}(X,t)$) in Equation \ref{kde}, $w_{k}$ is the weight value given to sub-window $k$, and $m$ is the number of sub-windows.
+\end{comment}
+
+****************begin from Sonja
+
+To address the second issue {\it Volatility of packet rates}, we ensure that we do not forget about the MAC addresses that were not observed in the last time window.
+In fact, for every MAC address that was ever observed, until it has been observed again we maintain the old probability distribution, too.
+Over time this probability distribution turns gradually into a uniform by smoothing it every epoch with a Gaussian kernel.
+The size of the kernel changes over time and is related to the maximal speed that the MAC address can have under the current density distribution of the crowd.
+In other words, we assume a Brownian motion of the MAC address, where the speed is limited by the maximal speed of a pedestrian under the current crowd conditions (cite paper Johansson et al).
+When the MAC address is observed again, its old probability distribution is simply overwritten by the probability distribution that is computed as above.
+
+Formally,  (now write formulas)
+
+****************end from Sonja
+
+***************begin from Sonja
+
+To address the last issue, {\it MAC address randomization}, we rely once again on the law of the large numbers. 
+*******end from Sonja
 
 \subsection{Validation}
-
-\begin{itemize}
-\item Simulation?\\
-Do we include simulation of the positioning process, i.e. the toy Monte Carlo simulation, or do we generate fitted positions with error values, and only test the density estimation methods?
-\end{itemize}
 
@@
 The relation between the number of App users and the crowd density is calibrated (using linear regression), which forms the basis of their participatory sensing method.
+**********from Sonja: but the main idea/proposition in this paper is that they are reverting the formula for speed as a function of density of Johansson, to infer the density when they know the speed (??)******end from sonja*****
 The accuracy of the smoothing method is assessed using ground truth crowd density information obtained from video footage recorded by CCTV cameras. 

### 2016-06-28 PaperOnSmoothing/draft-31052016.tex
Updated methods section.

@@
 
-The data provides us with a series of estimated positions, together with corresponding values for the uncertainty (see Section on Positioning).
-From the estimated positions we wish to generate density histograms.
+To visualise the spatiotemporal behaviour of crowd density [in real-time] we dynamically generate heat maps along a moving time interval. 
+The data provides us with a series of estimated positions of mobile devices, together with corresponding values of measurement uncertainty (see Section on Positioning).
+[The uncertainty values are determined by the co-variance matrix resulting from the chi-square fitting method.]
 
-To make density estimates we consider a two-dimensional binned region of space, and count the number of positions that fall into each bin.
-We wish to visualize / preserve the uncertainty for each estimated position.
-Therefore we transform each bin count into a bivariate Gaussian, using the separate error values ($\sigma_{x}$ and $\sigma_{y}$) for each dimension / direction.
-For each MAC we construct a two-dimensional density function that is the sum of these smooth positions.
-Because the number of measurements are different for each detected MAC address (in the interval) we normalize each histogram separately.
+From the data we select positions that are estimated from measurements that fall within a time window.
+From the estimated positions we construct crowd density estimations for the number of mobile devices per unit area.
+To visualize the uncertainty that corresponds to each estimated position, we transform it into a bi-variate Gaussian, using the uncertainty values ($\sigma_{x}$ and $\sigma_{y}$).
+For each MAC address detected we construct a two-dimensional density estimate that is the sum of these smoothed positions, which we normalize by the number of measurements $N$ for that MAC address.
+Thus, the final density estimate for a mobile device is a sum of `bumps' centered at the estimated positions.
+Because the number of measurements is different for each detected mobile device, we normalize each histogram separately.
+To build our final probability density estimate we sum all the density distributions of all the MAC addresses.
+This final density estimate integrates to the number of MAC addresses detected in the time window.
 
-We sum all the density histograms to create the total probability density histogram, which should ideally integrate to the number of MAC's detected (in the interval).
+The implementation of our method is similar to kernel density estimation \cite{scott}\cite{silverman}.
+In our case the amount of smoothing is variable and determined by the measurement uncertainty values $\sigma_{x}$ and $\sigma_{y}$.
 
-We smooth each position by evaluating the following function at each bin in the histogram,
+The density estimate for a mobile device $m$ at a location $(x,y)$ is defined by
 
 \begin{equation}
-\hat{d}(x,y)=K((x-x_{i}),\sigma_{x}) K((y-y_{i}),\sigma_{y})
+\hat{f}_{m}(x,y)=\frac{1}{N}\sum_{i=1}^{N}K((x-x_{i}),\sigma_{x}) K((y-y_{i}),\sigma_{y})
 \label{kde}
 \end{equation}
-where the location $(x,y)$ is expressed in bin unit, and where we use the kernel function
+where $N$ is the number of measurements (positions) for MAC address $m$, and the kernel function $K$ is give by
 \begin{equation}
@@
 \end{equation}
-Because the distances are expressed in bin units we scale the sigma's  ($\sigma_{x}$ and $\sigma_{y}$) by dividing the error values $\epsilon$ by the bin size $h$ 
+The final density estimate is given by
 \begin{equation}
-\sigma=\frac{\epsilon}{h}
+\hat{f}_{T}(x,y)=\sum_{m}\hat{f}_{m}(x,y)
+\label{total}
 \end{equation}
-This implies that the bin size becomes the unit for our density estimates, and we generate probabilities for number of people per bin, which has the area $h\times h$.
+To visualize our results as a density histogram in which the estimated values are probabilities for the number of people per bin, we evaluate equation \ref{total} at regular points in a grid, and multiply by the bin size.
 

### 2016-06-28 PaperOnSmoothing/draft-31052016.tex
MAC randomization in methods section

@@
 
-To address the last issue, {\it MAC address randomization}, we rely once again on the law of the large numbers. 
+To address the last issue, {\it MAC address randomization}, we rely once again on the fact that we have a lot of data.
+Namely, we know from the structure of the MAC address whether it is randomized or not.
+Figure \ref{fig:randomized} shows the ratio between randomized and non-randomized addresses observed per minute from midnight until 6:00 am.
+We can see that it is quite stable through time; in fact, the Pearson correlation coefficient between the time series of randomized and non-randomized addresses is 0.88.
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=130mm]{RatioRandNonrand.jpeg}
+	\caption{Estimations of x-coordinates of a static MAC device as a function of time}
+	\label{fig:randomized}
+\end{figure}
+
+Thus, when estimating crowd density, we ignore the MAC devices that have a randomized address and at the end we multiply the crowd density by 1.0225, to account for the ignored MAC devices.
+We choose factor 1.0225 to allow for a safety margin: in the peaks of the graph in Fig.
+\ref{fig:randomized} the values are 0.025, and we would rather overestimate than underestimate the crowd density.
+Note that this ratio should be re-computed periodically, to account for the changing conditions at the smart phones market.
+
 *******end from Sonja

### 2016-06-28 PaperOnSmoothing/draft-31052016.tex
MAC randomization in methods section + comments for Philip in methods section
corrections in MAC subsection of methods section

@@
 
-Formally,  (now write formulas)
+Formally,  (to Philip: now write formulas, please )
 
@@
 Thus, when estimating crowd density, we ignore the MAC devices that have a randomized address and at the end we multiply the crowd density by 1.0225, to account for the ignored MAC devices.
-We choose factor 1.0225 to allow for a safety margin: in the peaks of the graph in Fig.
-\ref{fig:randomized} the values are 0.025, and we would rather overestimate than underestimate the crowd density.
+We choose factor 1.225 to allow for a safety margin: in the peaks of the graph in Fig.
+\ref{fig:randomized} the values are 0.225, and we would rather overestimate than underestimate the crowd density.
 Note that this ratio should be re-computed periodically, to account for the changing conditions at the smart phones market.

### 2016-06-28 PaperOnSmoothing/draft-31052016.tex
pictures improved

@@
 
-The rest of the paper is organized as follows.....*******
-
-*************n.b.
-in related work section put all that use wi-fi tracking, for those that use GPS tracking only mention that this does not work in indoor spaces and that it requires participation.
-Not necessary to put work on video tracking in related work - it is already mentioned in intro.
-Mention also bluetooth tracking, but bluetooth is not so ubiquitous and our principles apply also if bluetooth was used (right?)
-Finally, none of the mentioned work uses our approach of modeling the position of an individual as a probability distribution; our method is designed to attack the problem of having a dense crowd; we are not interested in precise estimations for freely moving crowds; rather, we are interested in obtaining precise estimation when the  concert crowds is dense and static due to this.
-blah blah (sonja: explain better)********
+The rest of the paper is organized as follows.....
 
@@
 
-Sonja:Mention somewhere that in the documentary of the LP disaster people were raising their phones up to look for a way out. (which is good , less blockage of signals)
 \subsection{Issues with positioning (title to reconsider)}
@@
 			\centering
-			\includegraphics[width=100mm]{bimodal.jpg}
-			\caption{Trilateration can lead to multiple local optima (cut the lower part from the picture)}
+			\includegraphics[width=120mm]{bimodal.jpg}
+			\caption{Estimation of x-coordinate of a static MAC device through time with error}
 			\label{fig:bimodal}
@@
 	\includegraphics[width=130mm]{RatioRandNonrand.jpeg}
-	\caption{Estimations of x-coordinates of a static MAC device as a function of time}
+	\caption{Ratio between randomized and non-randomized addresses observed per minute}
 	\label{fig:randomized}
@@
 
+Sonja:Mention somewhere that in the documentary of the LP disaster people were raising their phones up to look for a way out. (which is good , less blockage of signals)
+
+*************n.b.
+in related work section put all that use wi-fi tracking, for those that use GPS tracking only mention that this does not work in indoor spaces and that it requires participation.
+Not necessary to put work on video tracking in related work - it is already mentioned in intro.
+Mention also bluetooth tracking, but bluetooth is not so ubiquitous and our principles apply also if bluetooth was used (right?)
+Finally, none of the mentioned work uses our approach of modeling the position of an individual as a probability distribution; our method is designed to attack the problem of having a dense crowd; we are not interested in precise estimations for freely moving crowds; rather, we are interested in obtaining precise estimation when the  concert crowds is dense and static due to this.
+blah blah (sonja: explain better)********
+
 \bibliography{references}

### 2016-06-29 PaperOnSmoothing/draft-31052016.tex
Updates on Method section.

@@
 \begin{comment}
-
 To visualise the spatiotemporal behaviour of crowd density in real-time we dynamically generate heat maps along a moving time interval.
@@
 The generation of dynamic heat maps allows us to detect and locate hot-spots of density where crowd conditions reach critical values that could possibly lead to a disaster.
-
-To construct two-dimensional probability density functions we apply non-parametric density estimation.
-For each bin location $X$, in the time interval $t$, we apply the kernel density estimation (KDE) method \cite{scott}\cite{silverman} given by
-
 \end{comment}
@@
 We don't generate probabilities for number of people per bin, but we generate an estimation of number of people per bin.
-}
+}\\
 
-To visualise the spatiotemporal behaviour of crowd density [in real-time] we dynamically generate heat maps along a moving time interval. 
-The data provides us with a series of estimated positions of mobile devices, together with corresponding values of measurement uncertainty (see Section on Positioning).
-[The uncertainty values are determined by the co-variance matrix resulting from the chi-square fitting method.]
+The data provides us with a series of estimated positions of mobile devices, together with their corresponding values of measurement uncertainty (see Section on Positioning).
+\begin{comment}
+The uncertainty values are determined by the co-variance matrix resulting from the chi-square fitting method.
+\end{comment}
+From this data we wish to estimate crowd density distributions, along a moving time window.
+So, our first step is to select from the data estimated positions whose time stamps fall within a specified time interval $[t-\Delta t,t]$.
+From the selected estimated positions we then construct density estimates for the individual mobile devices, and add them up to construct the crowd density distribution.
 
-From the data we select positions that are estimated from measurements that fall within a time window.
-From the estimated positions we construct crowd density estimations for the number of mobile devices per unit area.
-To visualize the uncertainty that corresponds to each estimated position, we transform it into a bi-variate Gaussian, using the uncertainty values ($\sigma_{x}$ and $\sigma_{y}$).
-For each MAC address detected we construct a two-dimensional density estimate that is the sum of these smoothed positions, which we normalize by the number of measurements $N$ for that MAC address.
-Thus, the final density estimate for a mobile device is a sum of `bumps' centered at the estimated positions.
-Because the number of measurements is different for each detected mobile device, we normalize each histogram separately.
-To build our final probability density estimate we sum all the density distributions of all the MAC addresses.
-This final density estimate integrates to the number of MAC addresses detected in the time window.
+A natural way to construct density estimates for the individual devices would be to construct density histograms, by binning the estimated positions.
+However, in constructing the density estimates we wish to preserve and visualize the uncertainty that corresponds to each estimated position.
+Therefore, we smooth each position into a bivariate Gaussian `bump', using the uncertainty values ($\sigma_{x}$ and $\sigma_{y}$) provided by the data.
+For each detected MAC address we construct a two-dimensional density estimate by adding up these smoothed positions, which we normalize by the number of measurements $N$ we have for that MAC address in the selected time window.
+Thus, the density estimate for a mobile device is the sum of a series of Gaussian `bumps' with variable width, centered at the estimated positions.
 
-The implementation of our method is similar to kernel density estimation \cite{scott}\cite{silverman}.
+Finally, to construct our crowd density estimation we sum the individual density estimates of all the devices.
+This estimated crowd density distribution integrates to the number of MAC addresses detected in the time window.
+
+The implementation of our method is similar to that of kernel density estimation \cite{scott}\cite{silverman}.
 In our case the amount of smoothing is variable and determined by the measurement uncertainty values $\sigma_{x}$ and $\sigma_{y}$.
-
 The density estimate for a mobile device $m$ at a location $(x,y)$ is defined by
@@
 \end{equation}
-where $N$ is the number of measurements (positions) for MAC address $m$, and the kernel function $K$ is give by
+where $N$ is the number of measurements (positions) for MAC address $m$, and the kernel function $K$ is given by
 \begin{equation}
@@
 \end{equation}
-The final density estimate is given by
+The final crowd density estimation is given by
 \begin{equation}
@@
 \end{equation}
-To visualize our results as a density histogram in which the estimated values are probabilities for the number of people per bin, we evaluate equation \ref{total} at regular points in a grid, and multiply by the bin size.
 
@@
 
-Formally,  (to Philip: now write formulas, please )
+Formally,  (now write formulas)
 
-****************end from Sonja
+****************end from Sonja\\
+
+\begin{comment}
+below is a first sketch, nothing definite yet
+\end{comment}
+
+For the epochs during which a mobile device is not detected we assume Brownian motion (without drift), which is defined by
+
+\begin{equation}
+\text{d}X(t)=\sqrt{2D}\text{d}W(t)
+\end{equation}
+where $W(t)$ is a Wiener process. The density function $f(t,x)$ of the displacement of $X$ in the time interval $[0,t]$ is known, and is a normal distribution with mean $X(0)$ and variance
+\begin{equation}
+\sigma^2=2Dt
+\end{equation}
+Thus, in case have no new observations, we can further smooth the density estimate from the previous epoch by convolution with a normal distribution. 
+The variance is determined by the length $t$ of the time interval, and the (average) walking speed of pedestrians.
+The walking speed in turn can be related to the local crowd density via the so-called fundamental diagram.\\
 

### 2016-06-29 PaperOnSmoothing/draft-31052016.tex
methods section updated (issue 1)

--- 2016-06-28 PaperOnSmoothing/draft-31052016.tex
@@
 Nevertheless, we can safely assume that for a regular crowd at a concert the average correlation tends to zero as the number of people increases; thus, the error of the estimation of the crowd density will in anyway vanish as the number of people tends to infinity.
-Formally, let R be an arbitrary region.
-(now I have to write some formulas, will continue on Monday)....
+
+Formally, let $R$ be an arbitrary region from the stadium. Let $\{mac_1, mac_2,...mac_n\}$ be all MAC devices detected at time $t$.  Let $X_i$ be a random variable defined by
+
+$$X_i = 
+\left\{
+\begin{array}{ll}
+1  & \mbox{if } mac_i \in R \\
+0 & \mbox{if } mac_i \notin R
+\end{array}
+\right.  $$
+
+ Denote by $X$ the total number of devices in $R$ detected at time $t$. Clearly, $X=\sum_{i=1}^{n} X_i$.
+Then $E(X)$, the expected value of $X$ is $$ E(X) = E(\sum_{i=1}^{n} X_i) = \sum_{i=1}^{n}E( X_i) = 
+\sum_{i=1}^{n}(1\cdot Prob(mac_i \in R) + 0 \cdot Prob(mac_i \notin R)) = \sum_{i=1}^{n} Prob(mac_i \in R) ,$$
+where by $ Prob(mac_i \in R) $ we denote the probability that $mac_i$ is in the region R at time $t$.
+
+Now about the variance.................
 
@@
 
-To address the second issue {\it Volatility of packet rates}, we ensure that we do not forget about the MAC addresses that were not observed in the last time window.
-In fact, for every MAC address that was ever observed, until it has been observed again we maintain the old probability distribution, too.
+To address the second issue {\it Volatility of packet rates}, we ensure that we do not forget about the MAC devices that were not observed in the last time window.
+In fact, for every MAC device that was ever observed, until it has been observed again we maintain the old probability distribution, too.
 Over time this probability distribution turns gradually into a uniform by smoothing it every epoch with a Gaussian kernel.
-The size of the kernel changes over time and is related to the maximal speed that the MAC address can have under the current density distribution of the crowd.
-In other words, we assume a Brownian motion of the MAC address, where the speed is limited by the maximal speed of a pedestrian under the current crowd conditions (cite paper Johansson et al).
-When the MAC address is observed again, its old probability distribution is simply overwritten by the probability distribution that is computed as above.
+The size of the kernel changes over time and is related to the maximal speed that the MAC device can have under the current density distribution of the crowd.
+In other words, we assume a Brownian motion of the MAC device, where the speed is limited by the maximal speed of a pedestrian under the current crowd conditions (cite paper Johansson et al).
+When the MAC device is observed again, its old probability distribution is simply overwritten by the probability distribution that is computed as above.
 
@@
 
+Future work: indclude map of the venue in the calculations, perhaps not both dark  regions in \ref{fig:trilateration_problem} are accessible by people!
+
 \bibliography{references}

### 2016-06-29 PaperOnSmoothing/draft-31052016.tex
subsections added, some corrections

@@
 
+\subsubsection{Issue 1: creating a statistical ensemble}
 *************from Sonja
@@
 
-Formally, let $R$ be an arbitrary region from the stadium. Let $\{mac_1, mac_2,...mac_n\}$ be all MAC devices detected at time $t$.  Let $X_i$ be a random variable defined by
+Formally, 
+let $\{mac_1, mac_2,...mac_n\}$ be all MAC devices detected at time $t$. Let $R$ be an arbitrary region from the stadium.  Let $X_i$ be a random variable defined by
 
@@
 
+**** below: in the kernel function there is a typo:  the negative sign in front of $u^2$ is forgotten
 ***************end from Sonja
@@
 \end{comment}
-
-{\it below: emphasize that we measure in a time window.
-The final result of stacking is not a probability density distribution but a crowd density distribution.
-We don't generate probabilities for number of people per bin, but we generate an estimation of number of people per bin.
-}\\
 
@@
 
+\subsubsection{Issue 2: Conservation of mass }
 ****************begin from Sonja
@@
 
+\subsubsection{Issue 3: using Big data again}
+
 ***************begin from Sonja
@@
 
-Thus, when estimating crowd density, we ignore the MAC devices that have a randomized address and at the end we multiply the crowd density by 1.0225, to account for the ignored MAC devices.
+Thus, when estimating crowd density, we ignore the MAC devices that have a randomized address and at the end we multiply the crowd density by 1.225, to account for the ignored MAC devices.
 We choose factor 1.225 to allow for a safety margin: in the peaks of the graph in Fig.
@@
 \subsection{Validation}
+Maybe with crowd sourcing images and videos from Sensation 2015 Amsterdam White
 

### 2016-06-30 PaperOnSmoothing/draft-31052016.tex
update
variance analysis added in section 3.2.1

@@
 This means that when the device switches from a probing to an online mode or vice versa, the MAC address suddenly changes and thus the device cannot be followed anymore.
-(n.b.
-here or later: there is a tag that reveals whether the address is randomized or not)
+It is worth noting, however, that the MAC address comes with a tag that reveals whether the address belongs to a certain domain (is non-randomized) or is randomized.
 \end{enumerate}
@@
 In order to deal with issue 1 from above, let us start with the following observation: we are not interested in the individual locations of the MAC devices, but rather in the density of the crowd; we would like especially to have more precise estimation when the density reaches a certain threshold, so that the crowd that is not in the dense region can be alarmed promptly to move away from the dense region.
-So, we are interested in estimating density accurately when the crowd is not in a state of movement (because if people are able to freely move, they are not in a danger).
 This situation, dense crowd, means that there is more blockage of signals from bodies, less precise estimation of distance from Wi-Fi based on signal strength, which means that the rings in Fig.\ref{fig:trilateration_problem} would be even thicker.
 This exaggerates the effect of "bimodal" or even multi-modal distribution of position estimates.
-Thus, in a dense crowd, all we can derive from the positioning method for a MAC address is a probability distribution of its possible locations.
+Thus, in a dense crowd, all we can derive from the positioning method for a MAC device is a probability distribution of its possible locations.
 For example, in the scenario in Fig.
 \ref{fig:trilateration_problem}, we can say that the MAC device is with a probability of 0.5 in the left grey region (region where all three rings overlap) and with a probability of 0.5 in the right grey region.
-While this does not provide us with useful information for the location of the MAC address, if we apply the same reasoning for all MAC addresses, and, for a region R we add together the probabilities for all MAC devices that they are in that region, we end up with an estimation of the number of MAC devices in that region.
-If we assume that the probability distributions of the locations of the MAC addresses are independent, we can directly apply the law of the large numbers and conclude that for a dense crowd the error of the estimation of the density per region will vanish.
+While this does not provide us with useful information for the location of the MAC device, if we apply the same reasoning for all MAC devices, and, for a region R we add together the probabilities for all MAC devices that they are in that region, we end up with an estimation of the number of MAC devices in that region.
+If we assume that the locations of the MAC devices are mutually independent, we can  apply directly the law of the large numbers and conclude that for a dense crowd the error of the estimation of the density per region will vanish.
 However, we cannot assume the mentioned independence, because people tend to go in groups to concerts.
-In this case, the variance of the estimation is equal to the average correlation between the distributions in the limiting case.
-Nevertheless, we can safely assume that for a regular crowd at a concert the average correlation tends to zero as the number of people increases; thus, the error of the estimation of the crowd density will in anyway vanish as the number of people tends to infinity.
+In this case, the variance of the estimation is equal to the average correlation between the locations in the limiting case.
+Nevertheless, we can safely assume that for a regular crowd at a concert the average correlation tends to zero as the number of people increases; thus, the error of the estimation of the crowd density will in anyway diminish as the number of people tends to infinity.
 
@@
 \sum_{i=1}^{n}(1\cdot Prob(mac_i \in R) + 0 \cdot Prob(mac_i \notin R)) = \sum_{i=1}^{n} Prob(mac_i \in R) ,$$
-where by $ Prob(mac_i \in R) $ we denote the probability that $mac_i$ is in the region R at time $t$.
+where by $Prob(mac_i \in R)$ we denote the probability that $mac_i$ is in the region R at time $t$.
+We postpone the derivation of $Prob(mac_i \in R)$ a bit.
 
-Now about the variance.................
+Next, we show that the variance of $X/n$, that is, the number of devices in $R$ normalized w.r.t the total number of detected devices, diminishes when $n$ becomes large (the variance of $X$ in the limiting case is out of our interest because then $E(X)$ is potentially infinite).
+This suffices to show that our method of estimation of crowd density is theoretically sound, given the probabilities $Prob(mac_i \in R)$.
+We have
+
+\begin{equation}\label{eq_Var}
+Var(\frac{X}{n} ) =  Var(\frac{1}{n} \cdot \sum_{i=1}^{n} X_i) = \frac{1}{n^2} \cdot Var(\sum_{i=1}^{n} X_i) = \frac{1}{n^2} \cdot (\sum_{i=1}^{n} Var(X_i) + \sum_{i \not =  j}Cov(X_i,X_j))
+\end{equation}
+
+We can assume that there is an upper limit $\lambda$ of a number of people that go to a concert together (a group), that is, whose locations are correlated\footnote{Actually, when the crowd is so dense that people can not move freely anymore, the whole crowd becomes a "group" (cite LP disaster) and all locations are correlated.
+But we aim to detect a high density with our method {\it before} this happens, to react preventively; otherwise, it is too late.}.
+Denote by $\kappa$ the maximal covariance between any $X_i$ and $X_j$ and let us write $i \sim j$ if and only if $i$ and $j$ are in the same group.
+Then, 
+\begin{equation}\label{eq_covar}
+\sum_{i \not =  j}Cov(X_i,X_j) = 2 \sum_{1\leq i <  j \leq n }Cov(X_i,X_j) =  2 \sum_{i,j: i \sim j }Cov(X_i,X_j)  \leq 2 n \cdot \frac{\lambda(\lambda - 1)}{2} \cdot\kappa = n  k \lambda(\lambda - 1)
+\end{equation} 
+where the inequality holds because the maximal number of groups is $n$ and the maximal number of pairs $(i,j)$ in a group is $\lambda(\lambda-1)/2$.
+Let $\mu = \frac{1}{n}\sum_{i=1}^{n} Var(X_i)$, that is, 
+denote by $\mu$ the average variance of $X_i$.  From \eqref{eq_Var} and \eqref{eq_covar} we have
+
+\begin{equation}
+Var(\frac{X}{n} ) \leq \frac{1}{n^2}(n \mu + n  k \lambda(\lambda - 1)) = \frac{1}{n}(\mu +  k \lambda(\lambda - 1)), 
+\end{equation}
+which tends to $0$ when $n \rightarrow \infty$. Note that we have greatly overestimated the variance with the inequation in \eqref{eq_covar}.
+
+Next, we can proceed with derivation of $Prob(mac_i) \in R$ for arbitrary $i$ in order to evaluate $E(X)$.
 
@@
 Note that this ratio should be re-computed periodically, to account for the changing conditions at the smart phones market. 
-
+********mention the absolute numbers, or give graphs)
 *******end from Sonja
@@
 
-Schauer \textit{et al.} (2014) \cite{schauer:1} count unique MAC addresses detected by two sensors (nodes) at both sides (public and security) of a security check inside a major German airport, to estimate pedestrian densities and pedestrian flow.
+Schauer \textit{et al.} (2014) \cite{schauer:1} count unique MAC devices detected by two sensors (nodes) at both sides (public and security) of a security check inside a major German airport, to estimate pedestrian densities and pedestrian flow.
 They consider time information, to determine the direction of a person's movement, and at least one RSSI value, to reduce the number of false positives in case devices are captured by both sensors.

### 2016-07-04 PaperOnSmoothing/draft-31052016.tex
typos corrected
Continued a bit on the Methods section.

@@
 \begin{equation}\label{eq_covar}
-\sum_{i \not =  j}Cov(X_i,X_j) = 2 \sum_{1\leq i <  j \leq n }Cov(X_i,X_j) =  2 \sum_{i,j: i \sim j }Cov(X_i,X_j)  \leq 2 n \cdot \frac{\lambda(\lambda - 1)}{2} \cdot\kappa = n  k \lambda(\lambda - 1)
+\sum_{i \not =  j}Cov(X_i,X_j) = 2 \sum_{1\leq i <  j \leq n }Cov(X_i,X_j) =  2 \sum_{i,j: i \sim j }Cov(X_i,X_j)  \leq 2 n \cdot \frac{\lambda(\lambda - 1)}{2} \cdot\kappa = n  \kappa \lambda(\lambda - 1)
 \end{equation} 
@@
 \begin{equation}
-Var(\frac{X}{n} ) \leq \frac{1}{n^2}(n \mu + n  k \lambda(\lambda - 1)) = \frac{1}{n}(\mu +  k \lambda(\lambda - 1)), 
+Var(\frac{X}{n} ) \leq \frac{1}{n^2}(n \mu + n  \kappa \lambda(\lambda - 1)) = \frac{1}{n}(\mu +  \kappa \lambda(\lambda - 1)), 
 \end{equation}
@@
 \begin{equation}
-\text{d}X(t)=\sqrt{2D}\text{d}W(t)
+\text{d}X(t)=\sigma\text{d}W(t)
 \end{equation}
@@
 \begin{equation}
-\sigma^2=2Dt
+\text{Var(\textit{X(t)})}=\sigma^{2}t
 \end{equation}
-Thus, in case have no new observations, we can further smooth the density estimate from the previous epoch by convolution with a normal distribution. 
-The variance is determined by the length $t$ of the time interval, and the (average) walking speed of pedestrians.
-The walking speed in turn can be related to the local crowd density via the so-called fundamental diagram.\\
+Thus, in case we have no new observations, we smooth the density estimate from the previous epoch by convolution with the normal density function. 
+The variance is determined by the length $t$ of the time interval, and the walking speed of pedestrians. 
+[Optional: We relate $\sigma$ to the walking speed of pedestrians by setting $\sigma^2=2Dt$, where ]
+\begin{equation}
+D=\frac{(\Delta X)^2}{2\Delta t}
+\end{equation}
+The walking speed can be related to the local crowd density via the so-called \textit{Fundamental Diagram}. 
+When the density increases the movement of pedestrians becomes more impeded.
+The fundamental diagram has been the subject of extensive study [examples].
+Here we use the fundamental diagram as provided by ... [?]
+
+Smoothing a two-dimensional (input) function by convolution with a Gaussian weight function that adapts its width to the local values of the input function is [complicated and] time-consuming.\\
 

### 2016-07-07 PaperOnSmoothing/draft-31052016.tex
One or two revisions on Methods section.

@@
 The variance is determined by the length $t$ of the time interval, and the walking speed of pedestrians.
-[Optional: We relate $\sigma$ to the walking speed of pedestrians by setting $\sigma^2=2Dt$, where ]
+[Optional: We relate $\sigma^2$ to the walking speed of pedestrians as defined by]
 \begin{equation}
-D=\frac{(\Delta X)^2}{2\Delta t}
+\sigma^2=\frac{(\Delta X)^2}{\Delta t}
 \end{equation}
-The walking speed can be related to the local crowd density via the so-called \textit{Fundamental Diagram}. 
-When the density increases the movement of pedestrians becomes more impeded.
-The fundamental diagram has been the subject of extensive study [examples].
+The walking speed of pedestrians can be related to the local crowd density via the so-called \textit{Fundamental Diagram}. 
+The fundamental diagram generally states that walking speed of pedestrians is a decreasing function of the local crowd density [(Weidman 1992)].
+The fundamental diagram function has been the subject of extensive study [examples].
+Although all these studies agree on the decreasing form of the function, empirical studies in different countries and cultures have resulted in different forms and parameters values.
+
 Here we use the fundamental diagram as provided by ... [?]
-
-Smoothing a two-dimensional (input) function by convolution with a Gaussian weight function that adapts its width to the local values of the input function is [complicated and] time-consuming.\\
 

### 2016-07-07 PaperOnSmoothing/draft-31052016.tex
intro modified a bit

--- 2016-06-30 PaperOnSmoothing/draft-31052016.tex
@@
 \title{Crowd density estimation from Wi-Fi positioning data in the Amsterdam Arena}
+\author{Sonja Georgievska, Philip Rutten, Jan Amoraal, Michael Lees, Sander Klous\footnote{authors and order could change by the time the paper is submitted}}
 \maketitle
@@
 A lot of research on estimating crowd density has been done using video processing from security cameras (cite papers (1), (2), (3) from proposal?).
-However, this approach does not apply in our case, because 1)  it is difficult to obtain macroscopic overview ; 2) the lighting conditions during concert hours are not sufficient for video-based crowd analysis; and 3) the error of video-based density estimation increases with the increase of the actual crowd density.
+However, this approach does not suffice in our case, because 1)  it is difficult to obtain macroscopic overview ; 2) the lighting conditions during concert hours might not be sufficient for video-based crowd analysis; and 3) the error of video-based density estimation increases with the increase of the actual crowd density.
 
-In our approach, we combat the three above mentioned issues by exploiting the ubiquity of smart phones, as in ((4) (5), (6) from proposal?).
-Our living laboratory is the Amsterdam Arena(citation).
+In our approach, which can complement the video-based analysis, we combat the three above mentioned issues by exploiting the ubiquity of smart phones, as in ((4) (5), (6) from proposal?).
+Our living laboratory is the Amsterdam Arena(citation from proposal).
 Or methodology consists of three steps.
-First, we are using the Wi-Fi routers in the stadium to gather the strengths of the wireless antena signals from individuals smart phones.
-Second, we are estimating anonymously the positions of the visitors from the signal strengths in real time.
+First, we are using Wi-Fi routers in the stadium to gather the strengths of the wireless antena signals from individuals smart phones.
+Second, we are estimating anonymously the positions of the visitors from the signal strengths.
 This technique has been presented in (cite Jan's slides), and is similar to trilateration, that is, the way the Global Positioning System (GPS) is implemented.
-However, due to signal interference and blockage in dense crowds, irregular internet packet rates and systematic errors leading to multiple local optima, the positioning itself does not suffice to estimate the crowd density.
-Thus, in our third step, we aggregate the results from the positioning and apply the law of the large numbers to obtain an estimation of the crowd density, also in real-time.
-With the first two steps, we are dealing with issues 1) and 2) mentioned above.
+However, due to signal interference and blockage in dense crowds, irregular internet packet rates, and systematic errors leading to multiple local optima, the positioning itself does not suffice to estimate the crowd density.
+Thus, in our third step, we aggregate the results from the positioning and apply statistical ensembling to obtain an estimation of the crowd density.
+With the first two steps, we are handling issues 1) and 2) mentioned above.
 With the third step, which is the main contribution of this paper, we are handling issue 3) -  that is, with our method the precision of estimation actually increases as the crowd density increases.
@@
 
+Also future work: the speed in the brownian motion should depend on the local crowd density, rather than on the average crowd density.
+
 \bibliography{references}

### 2016-07-11 PaperOnSmoothing/draft-31052016.tex
references added

@@
 The Love Parade disaster (Duisburg, 2010), the Ellis Park Stadium disaster (Johannesburg, 2001), the PhilSports Stadium stampede (Manila, 2006) are just a few recent examples.
-One of the reasons that disasters happen is a "lack of overview of everybody" [cite paper on love parade EPJ DS], that is, a lack of macroscopic overview of the crowd.
-Critical crowd density (cite paper 8/m2, find it in e-mails)) is a contributing factor to crowd disasters; yet, it is still challenging to determine timely when it occurs, so that a disaster can be prevented before it happens, by e.g.
+One of the reasons that disasters happen is a "lack of overview of everybody" \cite{Helbing2012}, that is, a lack of macroscopic overview of the crowd.
+Critical crowd density  \cite{Helbing2012, density} is a contributing factor to crowd disasters; yet, it is still challenging to determine timely when it occurs, so that a disaster can be prevented before it happens, by e.g.
 navigating the rest of the crowd away from the congestion.
@@
 In our work we are interested in estimating crowd density during concerts in indoor spaces.
-A lot of research on estimating crowd density has been done using video processing from security cameras (cite papers (1), (2), (3) from proposal?).
+A lot of research on estimating crowd density has been done using video processing from security cameras (\cite{Crowded_Scene_Analysis, Krausz2012307}).
 However, this approach does not suffice in our case, because 1)  it is difficult to obtain macroscopic overview ; 2) the lighting conditions during concert hours might not be sufficient for video-based crowd analysis; and 3) the error of video-based density estimation increases with the increase of the actual crowd density.
@@
 In our approach, which can complement the video-based analysis, we combat the three above mentioned issues by exploiting the ubiquity of smart phones, as in ((4) (5), (6) from proposal?).
-Our living laboratory is the Amsterdam Arena(citation from proposal).
-Or methodology consists of three steps.
+Our living laboratory is the Amsterdam Arena \cite{arena}.
+Our methodology consists of three steps.
 First, we are using Wi-Fi routers in the stadium to gather the strengths of the wireless antena signals from individuals smart phones.
 Second, we are estimating anonymously the positions of the visitors from the signal strengths.
-This technique has been presented in (cite Jan's slides), and is similar to trilateration, that is, the way the Global Positioning System (GPS) is implemented.
+This technique has been presented in \cite{customers_Jan}, and is similar to trilateration, that is, the way the Global Positioning System (GPS) is implemented.
 However, due to signal interference and blockage in dense crowds, irregular internet packet rates, and systematic errors leading to multiple local optima, the positioning itself does not suffice to estimate the crowd density.
@@
 \begin{equation}
-\sigma^2=2Dt
+\text{Var(\textit{X(t)})}=\sigma^{2}t
 \end{equation}
-Thus, in case have no new observations, we can further smooth the density estimate from the previous epoch by convolution with a normal distribution. 
-The variance is determined by the length $t$ of the time interval, and the (average) walking speed of pedestrians.
-The walking speed in turn can be related to the local crowd density via the so-called fundamental diagram.\\
+Thus, in case we have no new observations, we smooth the density estimate from the previous epoch by convolution with the normal density function. 
+The variance is determined by the length $t$ of the time interval, and the walking speed of pedestrians.
+[Optional: We relate $\sigma^2$ to the walking speed of pedestrians as defined by]
+\begin{equation}
+\sigma^2=\frac{(\Delta X)^2}{\Delta t}
+\end{equation}
+The walking speed of pedestrians can be related to the local crowd density via the so-called \textit{Fundamental Diagram}. 
+The fundamental diagram generally states that walking speed of pedestrians is a decreasing function of the local crowd density [(Weidman 1992)].
+The fundamental diagram function has been the subject of extensive study [examples].
+Although all these studies agree on the decreasing form of the function, empirical studies in different countries and cultures have resulted in different forms and parameters values.
+
+Here we use the fundamental diagram as provided by ... [?]
 
@@
 
-Future work: indclude map of the venue in the calculations, perhaps not both dark  regions in \ref{fig:trilateration_problem} are accessible by people!
+Future work: include map of the venue in the calculations, perhaps not both dark  regions in \ref{fig:trilateration_problem} are accessible by people!
 

### 2016-07-12 PaperOnSmoothing/draft-31052016.tex
abstract added

@@
 \begin{abstract}
+
+We present a new methodology for estimating crowd density during festivals.
+Our motivation comes from the fact that many crowd disasters have happened because of abnormal crowd density.
+We exploit the ubiquity of smart phones, without requiring participation from the crowd, and we use Amsterdam ArenA as a living laboratory.
+We build upon previous work, where the location of a visitor is estimated anonymously based on the strengths of the Wi-Fi signals received at the ArenA Wi-Fi routers,  by a method similar to trilateration,.
+However, because in dense crowds a Wi-Fi signal is partially absorbed and partially reflected by the human bodies, the localization based on signals strengths alone does not suffice to estimate the crowd density.
+Moreover, since we don't rely on the smart phones being connected to the Wi-Fi network, the signal rates are quite volatile, and the MAC addresses of the phones may change, depending on the vendor's privacy policy.
+In this paper, we address those issues.
+We borrow ideas from statistical mechanics to estimate the crowd density: rather than estimating the position of a visitor, we create a continuous probability distribution over all possible positions.
+We rely on the law of the large numbers, with adaptation to correlated variables, to estimate the crowd density by aggregating the individual distributions.
+The individual probability distribution turns gradually into a uniform distribution, with which we address the problem of outdated signals.
+Finally, we use the law of large numbers again and the "randomized address" tag to account for the fact that a portion of the MAC addresses are randomized.
+Thus, the advantage of our methodology when compared to similar methodologies is that our estimation becomes more (rather than less) precise when the number of people increases.
+This suits the purpose of being able to detect raised crowd density.
+We validate our methodology by comparing it to the ground truth obtained from security cameras. \marginpar{val. to be done}
+
 \begin{comment}
@@
 
-***********************************From Sonja. nb: the text will be polished in the final phases, first I care only about putting all ideas on paper***************
+\marginpar{begin from Sonja. nb: the text will be polished in the final phases, first I care only about putting all ideas on paper}
 
@@
 
-In our approach, which can complement the video-based analysis, we combat the three above mentioned issues by exploiting the ubiquity of smart phones, as in ((4) (5), (6) from proposal?).
+In our approach, which can complement the video-based analysis, we combat the three above mentioned issues by exploiting the ubiquity of smart phones.
 Our living laboratory is the Amsterdam Arena \cite{arena}.
@@
 
-************************************end from Sonja
+\marginpar{end from Sonja}
 \begin{comment}

### 2016-07-12 PaperOnSmoothing/draft-31052016.tex
introduction changed

@@
 We exploit the ubiquity of smart phones, without requiring participation from the crowd, and we use Amsterdam ArenA as a living laboratory.
-We build upon previous work, where the location of a visitor is estimated anonymously based on the strengths of the Wi-Fi signals received at the ArenA Wi-Fi routers,  by a method similar to trilateration,.
+We build upon previous work, where the location of a visitor is estimated anonymously based on the strengths of the Wi-Fi signals received at the ArenA Wi-Fi routers,  by a method similar to trilateration.
 However, because in dense crowds a Wi-Fi signal is partially absorbed and partially reflected by the human bodies, the localization based on signals strengths alone does not suffice to estimate the crowd density.
@@
 In this paper, we address those issues.
-We borrow ideas from statistical mechanics to estimate the crowd density: rather than estimating the position of a visitor, we create a continuous probability distribution over all possible positions.
+We borrow ideas from statistical mechanics to estimate the crowd density.
+\begin{comment} rather than estimating the position of a visitor, we create a continuous probability distribution over all possible positions.
 We rely on the law of the large numbers, with adaptation to correlated variables, to estimate the crowd density by aggregating the individual distributions.
@@
 Finally, we use the law of large numbers again and the "randomized address" tag to account for the fact that a portion of the MAC addresses are randomized.
-Thus, the advantage of our methodology when compared to similar methodologies is that our estimation becomes more (rather than less) precise when the number of people increases.
+Thus, \end{comment}
+The advantage of our methodology when compared to similar methodologies is that our estimation becomes more (rather than less) precise when the number of people increases.
 This suits the purpose of being able to detect raised crowd density.
-We validate our methodology by comparing it to the ground truth obtained from security cameras. \marginpar{val. to be done}
+We also validate our methodology experimentally by comparing the crowd density estimation to the ground truth obtained from security cameras. \marginpar{val. to be done}
 
-\begin{comment}
+\begin{comment} From Philip:
 We present the Amsterdam Arena project, which involves observing and managing crowd behaviour, using the Amsterdam Arena stadium as a living laboratory.
@@
 
-\marginpar{begin from Sonja. nb: the text will be polished in the final phases, first I care only about putting all ideas on paper}
-
 Crowd disasters have taken many human lives.
@@
 One of the reasons that disasters happen is a "lack of overview of everybody" \cite{Helbing2012}, that is, a lack of macroscopic overview of the crowd.
-Critical crowd density  \cite{Helbing2012, density} is a contributing factor to crowd disasters; yet, it is still challenging to determine timely when it occurs, so that a disaster can be prevented before it happens, by e.g.
-navigating the rest of the crowd away from the congestion.
+Critical crowd density  \cite{Helbing2012, density} is a contributing factor to crowd disasters.
+Yet, it is still challenging to determine timely when it occurs, so that a disaster can be prevented by navigating the rest of the crowd away from the congestion.
 
-In our work we are interested in estimating crowd density during concerts in indoor spaces.
-A lot of research on estimating crowd density has been done using video processing from security cameras (\cite{Crowded_Scene_Analysis, Krausz2012307}).
-However, this approach does not suffice in our case, because 1)  it is difficult to obtain macroscopic overview ; 2) the lighting conditions during concert hours might not be sufficient for video-based crowd analysis; and 3) the error of video-based density estimation increases with the increase of the actual crowd density.
+In our work we estimate crowd density during concerts in indoor spaces.
+A lot of research on estimating crowd density has been done using video processing from security cameras \cite{Crowded_Scene_Analysis, Krausz2012307}.
+However, this approach does not suffice to detect raised crowd density.
+First, it is difficult to obtain macroscopic overview of the crowd.
+Second,
+the lighting conditions during concert hours might not be sufficient for video-based crowd analysis only.
+Finally, the error of video-based density estimation increases with the increase of the actual crowd density.
 
-In our approach, which can complement the video-based analysis, we combat the three above mentioned issues by exploiting the ubiquity of smart phones.
+In our approach, which can complement the video-based analysis, we combat the three above mentioned issues by exploiting the ubiquity of smart phones, as has been done in (citations).
+Contrarily to most s
+uch approaches (citations), we don't require participation from the crowd.
 Our living laboratory is the Amsterdam Arena \cite{arena}.
+We build upon previous work \cite{customers_jan}, where the location of a visitor is estimated anonymously based on the strengths of the Wi-Fi signals received at the ArenA Wi-Fi routers,  by a method similar to trilateration (the way the Global Positioning System is implemented).
+However, because in dense crowds a Wi-Fi signal is partially absorbed and partially reflected by the human bodies, the localization based on signals strengths alone does not suffice to estimate the crowd density; e.g.
+multiple optimal position estimations are possible.
+Moreover, since we don't rely on the smart phones being connected to the Wi-Fi network, the signal rates are quite volatile, and the MAC addresses of the phones may change (be "randomized"), complying to some  vendors' privacy policy.
+In this paper, we address those issues.
+We borrow ideas from statistical mechanics to estimate the crowd density: rather than estimating the position of a visitor, we create a continuous probability distribution over all possible positions (which is usually bimodal).
+We rely on the law of the large numbers, with adaptation to correlated variables, to estimate the crowd density by aggregating the individual distributions.
+The individual probability distribution turns gradually into a uniform distribution, with which we address the problem of outdated signals.
+Finally, we use the law of large numbers again and the "randomized MAC address" tag, that is available with the data, to account for the fact that a portion of the MAC addresses are randomized.
+Thus, the advantage of our methodology when compared to other methodologies is that our estimation becomes more (rather than less) precise when the number of people increases.
+This suits the purpose of being able to detect raised crowd density.
+We also validate our methodology experimentally by comparing the crowd density estimation time series to the ground truth obtained from security cameras. \marginpar{val. to be done}
+
+\begin{comment}
 Our methodology consists of three steps.
@@
 Thus, in our third step, we aggregate the results from the positioning and apply statistical ensembling to obtain an estimation of the crowd density.
+
 With the first two steps, we are handling issues 1) and 2) mentioned above.
 With the third step, which is the main contribution of this paper, we are handling issue 3) -  that is, with our method the precision of estimation actually increases as the crowd density increases.
+\end{comment}
 
-The rest of the paper is organized as follows.....
+The rest of the paper is organized as follows. In Sec \ref{sec:positioning}
 
@@
 \section{Positioning of visitors using Wi-Fi sensors and smart phones}
+\label{sec:positioning}
 
@@
 \section{From positioning to estimation of density}
+\label{sec:density}
 

### 2016-07-12 PaperOnSmoothing/draft-31052016.tex
title changed, text improved a bit

@@
 
-\title{Crowd density estimation from Wi-Fi positioning data in the Amsterdam Arena}
+\title{Towards detecting critical crowd density with Wi-Fi positioning: Case Amsterdam ArenA}
 \author{Sonja Georgievska, Philip Rutten, Jan Amoraal, Michael Lees, Sander Klous\footnote{authors and order could change by the time the paper is submitted}}
@@
 In our approach, which can complement the video-based analysis, we combat the three above mentioned issues by exploiting the ubiquity of smart phones, as has been done in (citations).
-Contrarily to most s
-uch approaches (citations), we don't require participation from the crowd.
+Contrarily to most such approaches (citations), we \marginpar{add missing citations twice} don't require participation from the crowd.
 Our living laboratory is the Amsterdam Arena \cite{arena}.
@@
 This suits the purpose of being able to detect raised crowd density.
-We also validate our methodology experimentally by comparing the crowd density estimation time series to the ground truth obtained from security cameras. \marginpar{val. to be done}
+We also validate our methodology experimentally by comparing the crowd density estimation time series to the ground truth obtained from security cameras. \marginpar{validation to be done}
 
@@
 
-The rest of the paper is organized as follows. In Sec \ref{sec:positioning}
+The rest of the paper is organized as follows.
+In Sec.
+\ref{sec:positioning} we explain the process of trilateration, that estimates the position of a Wi-Fi device based on the strength of the signals captured by the Wi-Fi routers in the stadium.
+In Sec.
+\ref{sec:density} we present our methodology for estimating in real-time the spatial density distribution of the crowd.
+We first identify which issues have not been resolved by the positioning in order to estimate crowd density, and which are not related to the mathematical methodology behind the positioning step.
+Then, we present our approach towards resolving the issues.
+In Sec.
+\ref{sec:validation} we compare our estimation of density to the one  obtained by video analysis.
+In Sec.
+\ref{sec:relatedwork} we put our approach into a perspective by comparing it to other approaches that are using the smart phones.
+We conclude and point to future research directions in Sec.
+\ref{sec:conclusion}.
 
 \marginpar{end from Sonja}
-\begin{comment}
-\begin{itemize}
-
-\item Background\\
-
-Human crowds are complex systems, and predicting or controlling their behavior is challenging.
-Our approach involves three phases, firstly data collection from Wi-Fi and Bluetooth sensors in the stadium, secondly data analytics, and finally we aim to use simulation to make forecasts about the crowd dynamics.
-
-In this paper we present initial results on the data analytics and show how we can extract density maps from Wi-Fi positioning data.
-
-\item Aims\\
-
-The main purpose is to be able to predict a possible crowd disaster and identify means to prevent it from happening.
-
-\end{itemize}
-\end{comment}
 
@@
 
-Under ideal circumstances, the positioning itself would suffice to estimate the crowd density per region of interest:  every few seconds we would only need to count the number of estimates that fall in the region of interest.
+Under ideal circumstances, the positioning itself would suffice to estimate the spatial crowd density distribution:  every second we would only need to count the number of estimates per square meter.
 However, we have observed several issues, which are not related to the mathematical methodology behind the positioning step, that prevent us to apply direct counting.
@@
 	{\it Bimodal distributions of coordinates estimations}.
-We have sampled randomly 20 MAC addresses and plotted the estimations of their coordinates through time, where we have plotted only the estimations with a relatively small uncertainty.
+We have sampled randomly 20 MAC addresses and plotted the estimations of their coordinates through time, where we have plotted only the estimations with a relatively small (conditional) uncertainty.
 We observed persistent bimodal distribution of the estimations, an example of which can be seen in Fig.
@@
 \subsubsection{Issue 1: creating a statistical ensemble}
-*************from Sonja
+*************start from Sonja
 
@@
 where the inequality holds because the maximal number of groups is $n$ and the maximal number of pairs $(i,j)$ in a group is $\lambda(\lambda-1)/2$.
-Let $\mu = \frac{1}{n}\sum_{i=1}^{n} Var(X_i)$, that is, 
+Let $$\mu = \frac{1}{n}\sum_{i=1}^{n} Var(X_i),$$ that is, 
 denote by $\mu$ the average variance of $X_i$.  From \eqref{eq_Var} and \eqref{eq_covar} we have
@@
 \end{equation}
-which tends to $0$ when $n \rightarrow \infty$. Note that we have greatly overestimated the variance with the inequation in \eqref{eq_covar}.
+which tends to $0$ when $n \rightarrow \infty$.
+Note that we have greatly overestimated the covariance with the inequation in \eqref{eq_covar}, which means that in practice the variance converges to $0$ much faster than as presented.
 
@@
 
-**** below: in the kernel function there is a typo:  the negative sign in front of $u^2$ is forgotten
 ***************end from Sonja
@@
 \begin{equation}
-K(u,\sigma)=\frac{1}{\sigma\sqrt{2\pi}}\exp(\frac{u^2}{2\sigma^2})
+K(u,\sigma)=\frac{1}{\sigma\sqrt{2\pi}}\exp(-\frac{u^2}{2\sigma^2})
 \end{equation}
@@
 When the MAC device is observed again, its old probability distribution is simply overwritten by the probability distribution that is computed as above.
+\marginpar{insert a picture}
 
@@
 
-\subsection{Validation}
-Maybe with crowd sourcing images and videos from Sensation 2015 Amsterdam White
+\section{Validation}\label{sec:validation}
+Either with official videos from the organizational team of  Sensation 2015 Amsterdam White, or with security camera videos from Sensation 2016, in which case the analysis needs to be re-applied.
 
-\section{Related work}
+\section{Related work}\label{sec:relatedwork}
+****In final version: first split all into those that ask for participation and those that do not. Then, among those that do not ask for participation, say why our approach is better***
 
@@
 The accuracy of the smoothing method is assessed using ground truth crowd density information obtained from video footage recorded by CCTV cameras.
+
 The influence of the kernel radius on the correlation between the actual density and the density of App is analysed.
@@
 
-\section{Conclusion}
-
-Sonja:Mention somewhere that in the documentary of the LP disaster people were raising their phones up to look for a way out. (which is good , less blockage of signals)
-
 *************n.b.
@@
 Not necessary to put work on video tracking in related work - it is already mentioned in intro.
-Mention also bluetooth tracking, but bluetooth is not so ubiquitous and our principles apply also if bluetooth was used (right?)
-Finally, none of the mentioned work uses our approach of modeling the position of an individual as a probability distribution; our method is designed to attack the problem of having a dense crowd; we are not interested in precise estimations for freely moving crowds; rather, we are interested in obtaining precise estimation when the  concert crowds is dense and static due to this.
+Mention also bluetooth tracking, but bluetooth is not so ubiquitous and our principles apply also if bluetooth was used (right?) Finally, none of the mentioned work uses our approach of modeling the position of an individual as a probability distribution; our method is designed to attack the problem of having a dense crowd; we are not interested in precise estimations for freely moving crowds; rather, we are interested in obtaining precise estimation when the  concert crowds is dense and static due to this.
 blah blah (sonja: explain better)********
+
+\section{Conclusion}\label{sec:conclusion}
+
+Sonja:Mention somewhere that in the documentary of the LP disaster people were raising their phones up to look for a way out. (which is good , less blockage of signals)
 

### 2016-07-13 PaperOnSmoothing/draft-31052016.tex
text improved, synchronizing parts of sonja and philip on methodology

@@
 In this paper, we address those issues.
-We borrow ideas from statistical mechanics to estimate the crowd density: rather than estimating the position of a visitor, we create a continuous probability distribution over all possible positions (which is usually bimodal).
+We borrow ideas from statistical mechanics to estimate the crowd density: rather than estimating the most likely position of a visitor in real time, we create an evolving continuous probability distribution over all possible positions (which turns out to be usually bimodal).
 We rely on the law of the large numbers, with adaptation to correlated variables, to estimate the crowd density by aggregating the individual distributions.
@@
 
-\subsection{Issues with positioning (title to reconsider)}
+\subsection{Positioning only is not enough}
 
@@
 	\item
-	{\it Bimodal distributions of coordinates estimations}.
+	{\it Issue 1: bimodal distributions of coordinates estimations}.
 We have sampled randomly 20 MAC addresses and plotted the estimations of their coordinates through time, where we have plotted only the estimations with a relatively small (conditional) uncertainty.
@@
 		\end{figure} 
-	This figure shows the estimated x-coordinates through time of a static, 24 hours persistent MAC address.
-Because we did not observe  bi-modal distributions in the signal strengths, the bi-modal distribution of the positions estimates can be explained by the multiple local optima situation.
-Namely, due to signal interference, when the optimization of the estimation of location of a certain MAC address is performed, there can be multiple local optima that are equally good candidates.
+	This figure shows the estimated x-coordinates in meters through time of a static MAC address that was persistent for 24 hours.
+Because we did not observe  bi-modal distributions in the signal strengths, the bi-modal distribution of the positions estimates can be explained by a "multiple local optima" situation.
+Namely, in a dense crowd, when optimizing the positioning of a MAC device, there can be multiple local optima that are equally good candidates.
 For example, consider Fig.
 \ref{fig:trilateration_problem} \footnote{Picture courtesy of http://math.stackexchange.com/questions/42537/trilateration-with-bounds}.
-In the center of every ring there is a Wi-Fi router, that has estimated signal strength to a certain MAC device with a certain error range.
+\marginpar{change the position of the label and put units (meters)}In the center of every ring there is a Wi-Fi router, that has estimated signal strength to the MAC device with a certain error range.
 The error range is represented by the thickness of the ring.
 Then, there are two possible regions which are equally good candidates for the positioning of the MAC device, and those regions are the two darkest regions where all three rings overlap.
-Note that the problem can arise regardless of the thickness of the rings or the number of Wi-Fi routers.
+Note that the problem can arise regardless of the thickness of the rings and the number of Wi-Fi routers.
 	\begin{figure}[h!]
@@
 
-\item {\it Volatility of packet rates}.
-When a MAC device is connected to the WI-Fi internet, it sends packets with a relatively stable rate.
-However, during concert hours, very few devices are actually connected to the internet, and most of our estimations of positions come from signals that the device sends in a "probing" mode.
-In this case the packet rate is quite volatile, ranging from a few seconds to a few minutes (some citation?).
-This means that when we make a snapshot of the MAC addresses visible in a certain moment, we are only detecting a small fraction of the MAC devices.
+\item {\it Issue 2: volatility of packet rates}.
+When a MAC device is connected to the Wi-Fi internet, it sends packets with a relatively stable and frequent rate.
+However, during concert hours, very few devices are actually connected to the internet.
+So, most of our estimations of positions come from signals that the device sends in a "probing" mode, i.e.
+while searching for a network.
+In this case the packet rate is quite volatile, ranging from a few seconds to a few minutes (citation?)\marginpar{add citation}.
+This means that when we make a snapshot of the MAC addresses visible in a certain moment, we are detecting a only a very small fraction of the MAC devices.
 
-\item {\it MAC randomization}.
-Due to ever increasing privacy concerns and possibly other business reasons, starting from 2014, the Apple I-phones have introduced randomization of the MAC address when the device is in a probing mode (not connected to internet) (citation?).
+\item {\it Issue 3: MAC address randomization}.
+Due to ever increasing privacy concerns and possibly other business reasons, starting from 2014, the Apple I-phones have introduced randomization of the MAC address when the device is in a probing mode (not connected to internet) (citation?)\marginpar{add citation}.
 This means that when the device switches from a probing to an online mode or vice versa, the MAC address suddenly changes and thus the device cannot be followed anymore.
-It is worth noting, however, that the MAC address comes with a tag that reveals whether the address belongs to a certain domain (is non-randomized) or is randomized.
+It is worth noting, however, that from the MAC address itself it can be determined whether the address is an original MAC address (non-randomized) or randomized.
 \end{enumerate}
 
-\subsection{Addressing the issues: real-time crowd density estimation}
+\subsection{Crowd density estimation}
 
-\subsubsection{Issue 1: creating a statistical ensemble}
-*************start from Sonja
+\subsubsection{Bimodal distributions of coordinates estimations: Creating statistical ensembles}
 
-In order to deal with issue 1 from above, let us start with the following observation: we are not interested in the individual locations of the MAC devices, but rather in the density of the crowd; we would like especially to have more precise estimation when the density reaches a certain threshold, so that the crowd that is not in the dense region can be alarmed promptly to move away from the dense region.
-This situation, dense crowd, means that there is more blockage of signals from bodies, less precise estimation of distance from Wi-Fi based on signal strength, which means that the rings in Fig.\ref{fig:trilateration_problem} would be even thicker.
+In order to deal with issue 1 from above, let us start with the following observation: we are not interested in the individual locations of the MAC devices, but rather in the density of the crowd.
+Also, we would  especially like to have more precise estimation when the density reaches a certain threshold, so that the crowd outside the dense region can be promptly navigated away from it.
+This situation (dense crowd), means that there is more blockage and absorption of signals from bodies.
+This leads to less precise estimation of the distance between a MAC device and a Wi-Fi router based on Wi-Fi signal strength, which means that the rings in e.g.
+Fig.\ref{fig:trilateration_problem} would be even thicker.
 This exaggerates the effect of "bimodal" or even multi-modal distribution of position estimates.
-Thus, in a dense crowd, all we can derive from the positioning method for a MAC device is a probability distribution of its possible locations.
+Thus, in a dense crowd, the best that we can derive from the positioning method for a MAC device is a probability distribution over all possible locations.
 For example, in the scenario in Fig.
-\ref{fig:trilateration_problem}, we can say that the MAC device is with a probability of 0.5 in the left grey region (region where all three rings overlap) and with a probability of 0.5 in the right grey region.
-While this does not provide us with useful information for the location of the MAC device, if we apply the same reasoning for all MAC devices, and, for a region R we add together the probabilities for all MAC devices that they are in that region, we end up with an estimation of the number of MAC devices in that region.
-If we assume that the locations of the MAC devices are mutually independent, we can  apply directly the law of the large numbers and conclude that for a dense crowd the error of the estimation of the density per region will vanish.
-However, we cannot assume the mentioned independence, because people tend to go in groups to concerts.
-In this case, the variance of the estimation is equal to the average correlation between the locations in the limiting case.
-Nevertheless, we can safely assume that for a regular crowd at a concert the average correlation tends to zero as the number of people increases; thus, the error of the estimation of the crowd density will in anyway diminish as the number of people tends to infinity.
+\ref{fig:trilateration_problem}, we can say that the MAC device is with a probability of 0.5 in the left grey region (region where all three rings overlap) and with a probability of 0.5 in the right grey region \footnote{In this example we assign the probabilities in a trivial way for the purpose of demonstrating our idea}.
+While this does not provide us with  very useful information about the location of the MAC device, if we apply the same reasoning for all MAC devices, and we add together the spatial probability density functions for all MAC devices, we end up with a spatial distribution of the crowd density.
+If we assume that the locations of the MAC devices are mutually independent, we can  apply directly the law of the large numbers and conclude that for a dense crowd the error of the estimation of the density per square meter will vanish.
+However, we cannot assume the mentioned independence, because people tend to go to concerts in groups.
+In this case, the variance of the estimation in the limiting case is equal to the average correlation between the locations.
+Nevertheless, we can safely assume that for a regular crowd at a concert the group size is relatively small compared to the whole crowd and thus the average correlation tends to zero as the number of people increases.
+Thus, the error of the estimation of the crowd density will in anyway diminish as the number of people grows.
 
-Formally, 
-let $\{mac_1, mac_2,...mac_n\}$ be all MAC devices detected at time $t$. Let $R$ be an arbitrary region from the stadium.  Let $X_i$ be a random variable defined by
+In what follows we formalize the discussion above. 
+Let $\{mac_1, mac_2,...mac_n\}$ be all MAC devices detected at time $t$. Let $R$ be an arbitrary region from the stadium.  Let $X_i$ be a random variable defined by
 
@@
 \right.  $$
-
 Denote by $X$ the total number of devices in $R$ detected at time $t$. Clearly, $X=\sum_{i=1}^{n} X_i$.
@@
 We postpone the derivation of $Prob(mac_i \in R)$ a bit.
-
-Next, we show that the variance of $X/n$, that is, the number of devices in $R$ normalized w.r.t the total number of detected devices, diminishes when $n$ becomes large (the variance of $X$ in the limiting case is out of our interest because then $E(X)$ is potentially infinite).
+Instead, we first show that the variance of $X/n$, that is, the variance of the proportion of devices detected in $R$ out of all detected devices, diminishes when $n$ becomes large (note that the variance of $X$ in the limiting case is out of our interest because then $E(X)$ is potentially infinite).
 This suffices to show that our method of estimation of crowd density is theoretically sound, given the probabilities $Prob(mac_i \in R)$.
@@
 Let $$\mu = \frac{1}{n}\sum_{i=1}^{n} Var(X_i),$$ that is, 
-denote by $\mu$ the average variance of $X_i$.  From \eqref{eq_Var} and \eqref{eq_covar} we have
+denote by $\mu$ the average variance of ${X_1, X_2, ..., X_n}$.  From \eqref{eq_Var} and \eqref{eq_covar} we have
 
@@
 
-Next, we can proceed with derivation of $Prob(mac_i) \in R$ for arbitrary $i$ in order to evaluate $E(X)$.
-
-***************end from Sonja
-
-***start from Philip
+Next, we can proceed with derivation of $Prob(mac_i) \in R$ for arbitrary $i$ in order to be able to evaluate $E(X)$, the estimated crowd density per region.
 
@@
 
-The data provides us with a series of estimated positions of mobile devices, together with their corresponding values of measurement uncertainty (see Section on Positioning).
+The data provides us with a series of estimated positions of mobile devices, together with their corresponding values of measurement uncertainty (see Sec.\ref{sec:positioning}).
 \begin{comment}
 The uncertainty values are determined by the co-variance matrix resulting from the chi-square fitting method.
+
 \end{comment}
-From this data we wish to estimate crowd density distributions, along a moving time window.
-So, our first step is to select from the data estimated positions whose time stamps fall within a specified time interval $[t-\Delta t,t]$.
-From the selected estimated positions we then construct density estimates for the individual mobile devices, and add them up to construct the crowd density distribution.
+From this data we wish to estimate the spatial probability distribution for an arbitrary MAC device $m$, along a moving time window.
+So, our first step is to select from the data the estimated $N$ positions whose time stamps fall within a specified time interval $[t-\Delta t,t]$.
+A natural way to construct a two-dimensional probability distribution would be to construct a histogram, by binning the positions and normalizing with $N$.
+However, we also wish to preserve the uncertainty that corresponds to each estimated position.
+Therefore, we first "smooth" each position into a bivariate Gaussian `bump', using the uncertainty values ($\sigma_{x}$ and $\sigma_{y}$) provided by the data as standard deviations.
+Then, for  $m$ we construct a two-dimensional probability density function (p.d.f.) by adding up these bumps, and  dividing the sum by $N$.
+For example,
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=130mm]{non-smoothed-histo.png}
+	\hfill 
+	\includegraphics[width=130mm]{smoothed-histo.png}
+	\caption{Smoothing a histogram of recent positions}
+	\label{fig:smoothing}
 
-A natural way to construct density estimates for the individual devices would be to construct density histograms, by binning the estimated positions.
-However, in constructing the density estimates we wish to preserve and visualize the uncertainty that corresponds to each estimated position.
-Therefore, we smooth each position into a bivariate Gaussian `bump', using the uncertainty values ($\sigma_{x}$ and $\sigma_{y}$) provided by the data.
-For each detected MAC address we construct a two-dimensional density estimate by adding up these smoothed positions, which we normalize by the number of measurements $N$ we have for that MAC address in the selected time window.
-Thus, the density estimate for a mobile device is the sum of a series of Gaussian `bumps' with variable width, centered at the estimated positions.
-
-Finally, to construct our crowd density estimation we sum the individual density estimates of all the devices.
-This estimated crowd density distribution integrates to the number of MAC addresses detected in the time window.
+\end{figure}
 
@@
 In our case the amount of smoothing is variable and determined by the measurement uncertainty values $\sigma_{x}$ and $\sigma_{y}$.
-The density estimate for a mobile device $m$ at a location $(x,y)$ is defined by
+The p.d.f.  for a mobile device with an address  $m$ at a location $(x,y)$ is defined by
 
@@
 \end{equation}
-where $N$ is the number of measurements (positions) for MAC address $m$, and the kernel function $K$ is given by
+where the kernel function $K$ is given by
 \begin{equation}
@@
 \end{equation}
+Note that so far we assumed that in every time window there is at least one estimate for every MAC address ever detected.
+In what follows we explain how we capture the cases when this assumption does not hold.
 
@@
 
-\subsubsection{Issue 2: Conservation of mass }
-****************begin from Sonja
+\subsubsection{Volatility of packet rates: Applying "conservation of mass" principle }
 
-To address the second issue {\it Volatility of packet rates}, we ensure that we do not forget about the MAC devices that were not observed in the last time window.
+To address the second issue, {\it Volatility of packet rates}, we ensure that we do not forget about the MAC devices that were not observed in the last time window.
 In fact, for every MAC device that was ever observed, until it has been observed again we maintain the old probability distribution, too.
@@
 Formally,  (now write formulas)
-
-****************end from Sonja\\
 
@@
 
-\subsubsection{Issue 3: using Big data again}
-
-***************begin from Sonja
+\subsubsection{MAC address randomization: using Big data again}
 
@@
 Note that this ratio should be re-computed periodically, to account for the changing conditions at the smart phones market. 
-********mention the absolute numbers, or give graphs)
-*******end from Sonja
+\marginpar{mention the absolute numbers, or give graphs}
 
 \section{Validation}\label{sec:validation}
-Either with official videos from the organizational team of  Sensation 2015 Amsterdam White, or with security camera videos from Sensation 2016, in which case the analysis needs to be re-applied.
+To be done.
+Hopefully with official videos from the organizational team of  Sensation 2015 Amsterdam White.
+Note that by validation we mean more of calibration.
+In general the approach is theoretically valid, which we show in Sec.\ref{sec:density}.2.1, but we are using parameters in the model that need to be calibrated.
+We need independent data sets (videos) for calibration and testing, to avoid overfitting.
+Also the paremeters should be as much as possible initially optimized from the Wi-Fi data itself and from theory.
 
 \section{Related work}\label{sec:relatedwork}
-****In final version: first split all into those that ask for participation and those that do not. Then, among those that do not ask for participation, say why our approach is better***
+In progress. 
+****In final version: first split all into those that ask for participation from the crowd	 and those that do not.
+Then, among those that do not ask for participation, say why our approach is better***
 
@@
 \section{Conclusion}\label{sec:conclusion}
-
+To be done. 
 Sonja:Mention somewhere that in the documentary of the LP disaster people were raising their phones up to look for a way out. (which is good , less blockage of signals)

### 2016-07-14 PaperOnSmoothing/draft-31052016.tex
text improved, pictures added

@@
 \usepackage{graphicx}
+
+\usepackage{subcaption}
 \begin{document}
@@
 
-We can assume that there is an upper limit $\lambda$ of a number of people that go to a concert together (a group), that is, whose locations are correlated\footnote{Actually, when the crowd is so dense that people can not move freely anymore, the whole crowd becomes a "group" (cite LP disaster) and all locations are correlated.
+We can assume that there is an upper limit $\lambda$ of a number of people that go to a concert together (a group), that is, whose locations are correlated\footnote{Actually, when the crowd is so dense that people can not move freely anymore, the whole crowd becomes a "group" (as in the  Love Parade disaster) and all locations are correlated.
 But we aim to detect a high density with our method {\it before} this happens, to react preventively; otherwise, it is too late.}.
@@
 The data provides us with a series of estimated positions of mobile devices, together with their corresponding values of measurement uncertainty (see Sec.\ref{sec:positioning}).
-\begin{comment}
-The uncertainty values are determined by the co-variance matrix resulting from the chi-square fitting method.
 
-\end{comment}
+%The uncertainty values are determined by the co-variance matrix resulting from the chi-square fitting method.
+
 From this data we wish to estimate the spatial probability distribution for an arbitrary MAC device $m$, along a moving time window.
@@
 However, we also wish to preserve the uncertainty that corresponds to each estimated position.
-Therefore, we first "smooth" each position into a bivariate Gaussian `bump', using the uncertainty values ($\sigma_{x}$ and $\sigma_{y}$) provided by the data as standard deviations.
-Then, for  $m$ we construct a two-dimensional probability density function (p.d.f.) by adding up these bumps, and  dividing the sum by $N$.
-For example,
+Therefore, we first "smooth" each position into a bivariate Gaussian distribution (`bump'), using the uncertainty values ($\sigma_{x}$ and $\sigma_{y}$) provided by the data as standard deviations.
+Then, for  $m$ we construct a two-dimensional probability density function (p.d.f.) by adding up these bumps, and  dividing the sum by $N$.\marginpar{ref.
+Figure, make viz.
+smooth}.
+(In Fig.
+\ref{fig:smoothing} we show the result of smoothing a histogram of the device discussed in Fig.\ref{fig:bimodal}.) %For example,
+
 \begin{figure}[h!]
 	\centering
-	\includegraphics[width=130mm]{non-smoothed-histo.png}
-	\hfill 
-	\includegraphics[width=130mm]{smoothed-histo.png}
-	\caption{Smoothing a histogram of recent positions}
+	\begin{subfigure}{.4\textwidth}
+		\centering
+		\includegraphics[width=1\linewidth]{non-smoothed-histo.png}
+		%\caption{1a}
+		%\label{fig:sfig1}
+	\end{subfigure}%
+	\begin{subfigure}{.4\textwidth}
+		\centering
+		\includegraphics[width=1\linewidth]{smoothed-histo.png}
+		%\caption{1b}
+		%\label{fig:sfig2}
+	\end{subfigure}
+	\caption{Smoothing a histogram with Gaussian kernels. Left: Original histogram. Right: The smoothed histogram.}
 	\label{fig:smoothing}
-
 \end{figure}
@@
 \end{equation}
+To finish our  discussion, in order to evaluate $Prob(mac_i) \in R$ we need to integrate $\hat{f}_{m}(x,y)$ for $(x,y) \in R$.
+
 Note that so far we assumed that in every time window there is at least one estimate for every MAC address ever detected.
@@
 
+(Needs to be finished.)
 To address the second issue, {\it Volatility of packet rates}, we ensure that we do not forget about the MAC devices that were not observed in the last time window.
 In fact, for every MAC device that was ever observed, until it has been observed again we maintain the old probability distribution, too.
-Over time this probability distribution turns gradually into a uniform by smoothing it every epoch with a Gaussian kernel.
+Over time this probability distribution turns gradually into a uniform by smoothing it every epoch with a Gaussian kernel (see Fig.
+\ref{fig:uniform} for an example).
 The size of the kernel changes over time and is related to the maximal speed that the MAC device can have under the current density distribution of the crowd.
 In other words, we assume a Brownian motion of the MAC device, where the speed is limited by the maximal speed of a pedestrian under the current crowd conditions (cite paper Johansson et al).
-When the MAC device is observed again, its old probability distribution is simply overwritten by the probability distribution that is computed as above.
-\marginpar{insert a picture}
+When the MAC device is observed again, its old probability distribution is simply overwritten by the probability distribution that is computed as above.\marginpar{ref.
+picture}
 
-Formally,  (now write formulas)
+%\begin{figure}[h!]
+%	\begin{subfigure}
+%	\centering
+%	\includegraphics[width=60mm]{brownian-smoothing-0002.png}
+%	\end{subfigure}
+%	\begin{subfigure}
+%	\centering
+%	\includegraphics[width=60mm]{brownian-smoothing-0004.png}
+%    \end{subfigure}
+%	\begin{subfigure}
+%	\centering	
+%	\includegraphics[width=60mm]{brownian-smoothing-0006.png}
+%	\end{subfigure}	
+%	\caption{Turning a distribution gradually into uniform}
+%	\label{fig:uniform}		
+%\end{figure}
 
-\begin{comment}
-below is a first sketch, nothing definite yet
-\end{comment}
+\begin{figure}[h!]
+	\begin{subfigure}{.3\textwidth}
+		\centering
+		\includegraphics[width=1\linewidth]{brownian-smoothing-0002.png}
+		%\caption{1a}
+		%\label{fig:sfig1}
+	\end{subfigure}%
+	\begin{subfigure}{.3\textwidth}
+		\centering
+		\includegraphics[width=1\linewidth]{brownian-smoothing-0004.png}
+		%\caption{1b}
+		%\label{fig:sfig2}
+	\end{subfigure}
+	\begin{subfigure}{.3\textwidth}
+		\centering
+		\includegraphics[width=1\linewidth]{brownian-smoothing-0006.png}
+		%\caption{1c}
+		%\label{fig:sfig3}
+	\end{subfigure}
+	\caption{Turning a distribution gradually into uniform}
+	\label{fig:uniform}
+\end{figure}
 
-For the epochs during which a mobile device is not detected we assume Brownian motion (without drift), which is defined by
-
+Formally, for the epochs during which a mobile device is not detected we assume Brownian motion (without drift), which is defined by
+% had to change X to B because I used X for the random variables above. Sorry.
 \begin{equation}
-\text{d}X(t)=\sigma\text{d}W(t)
+\text{d}B(t)=\sigma\text{d}W(t)
 \end{equation}
-where $W(t)$ is a Wiener process. The density function $f(t,x)$ of the displacement of $X$ in the time interval $[0,t]$ is known, and is a normal distribution with mean $X(0)$ and variance
+where $W(t)$ is a Wiener process. The density function $f(t,x)$ of the displacement of $X$ in the time interval $[0,t]$ is known, and is a normal distribution with mean $B(0)$ and variance
 \begin{equation}
-\text{Var(\textit{X(t)})}=\sigma^{2}t
+\text{Var(\textit{B(t)})}=\sigma^{2}t
 \end{equation}
@@
 \begin{equation}
-\sigma^2=\frac{(\Delta X)^2}{\Delta t}
+\sigma^2=\frac{(\Delta B)^2}{\Delta t}
 \end{equation}
@@
 Here we use the fundamental diagram as provided by ... [?]
+\marginpar{add rescaling to include information that all visitors are still inside the stadium }
 
@@
 To address the last issue, {\it MAC address randomization}, we rely once again on the fact that we have a lot of data.
-Namely, we know from the structure of the MAC address whether it is randomized or not.
+Namely, we know from the structure of the MAC address whether it has been randomized or not.
 Figure \ref{fig:randomized} shows the ratio between randomized and non-randomized addresses observed per minute from midnight until 6:00 am.
@@
 
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=130mm]{nonrandomized.jpeg}
+	\caption{Non-randomized addresses observed per minute}
+	\label{fig:nonrandomizedadresses}
+\end{figure} 
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=130mm]{randomized.jpeg}
+	\caption{Randomized addresses observed per minute}
+	\label{fig:randomizedaddresses}
+\end{figure} 
 \begin{figure}[h!]
@@
 
-Thus, when estimating crowd density, we ignore the MAC devices that have a randomized address and at the end we multiply the crowd density by 1.225, to account for the ignored MAC devices.
+\noindent Thus, when estimating crowd density, we ignore the MAC devices that have a randomized address and at the end we multiply the crowd density by 1.225, to account for the ignored MAC devices.
 We choose factor 1.225 to allow for a safety margin: in the peaks of the graph in Fig.

### 2016-07-15 PaperOnSmoothing/draft-31052016.tex
related work changed +  text changed

@@
 Therefore, we first "smooth" each position into a bivariate Gaussian distribution (`bump'), using the uncertainty values ($\sigma_{x}$ and $\sigma_{y}$) provided by the data as standard deviations.
-Then, for  $m$ we construct a two-dimensional probability density function (p.d.f.) by adding up these bumps, and  dividing the sum by $N$.\marginpar{ref.
-Figure, make viz.
-smooth}.
+Then, for  $m$ we construct a two-dimensional probability density function (p.d.f.) by adding up these bumps, and  dividing the sum by $N$.\marginpar{make viz.
+smooth and not dark}.
 (In Fig.
@@
 \ref{fig:randomized} the values are 0.225, and we would rather overestimate than underestimate the crowd density.
+\marginpar{in general: calculate it automatically, with a confidence interval}
 Note that this ratio should be re-computed periodically, to account for the changing conditions at the smart phones market. 
@@
 In progress.
-****In final version: first split all into those that ask for participation from the crowd	 and those that do not.
-Then, among those that do not ask for participation, say why our approach is better***
 
-Wirz \textit{et al.} (2012) \cite{wirz:1} infer crowd density from GPS location traces of people who use an App during the 2011 Lord Mayor Show in London.
+In the Introduction we mentioned the benefits of using smart phones based over video-based approach for density estimation of indoor concerts crowd; in fact, it is our opinion that the two approaches complement each other and in the future we plan to integrate both techniques in real time.
+Thus, in this section we are going to focus on previous work that estimates crowd density using wireless technologies.
+
+\cite{weppner:1} estimate crowd densities by distributing volunteers in the crowd, who are carrying smart phones scanning for Bluetooth devices.
+They then use statistics to combine the different measurements in space and time.
+They improve their method from mere counting to more advanced statistical analysis in \cite{weppner:2}, by using relative features that are more robust
+against statistical variations of the number of devices. The features include 
+the average speed of scanning devices, the average bluetooth signal strength and its variance, etc.
+
+%Wirz \textit{et al.} (2012) \cite{wirz:1} infer crowd density from GPS location traces of people who use an App during the 2011 Lord Mayor Show in London.
 They visualize the data in real-time, and generate heat maps (dynamically) using the kernel density estimation (KDE) method.
-In (Wirz \textit{et al.} 2013) \cite{wirz:2} the methodology is presented more elaborately, using a real-world data set collected during the same festival.
-The relation between the number of App users and the crowd density is calibrated (using linear regression), which forms the basis of their participatory sensing method.
-**********from Sonja: but the main idea/proposition in this paper is that they are reverting the formula for speed as a function of density of Johansson, to infer the density when they know the speed (??)******end from sonja*****
-The accuracy of the smoothing method is assessed using ground truth crowd density information obtained from video footage recorded by CCTV cameras.
+In  \cite{wirz:2} the authors follow a participatory sensing approach in which pedestrians share their locations on
+a voluntary basis. Since  only a fraction of
+all pedestrians share location information, they present a methodology to infer the crowd density from their walking speed, by inverting the formula for estimating maximal pedestrian speed given a crowd density, presented in \cite{Johannsson}.
+Note that in our case study the visitors of the concerts with a stage are mostly static without intention to reach a destination, so this technique cannot be directly applied.
 
-The influence of the kernel radius on the correlation between the actual density and the density of App is analysed.
-
-The Gaussian weight function used in (Wirz \textit{et al.} 2012; 2013) \cite{wirz:1}\cite{wirz:2} is introduced in Helbing \textit{et al.} (2007) \cite{helbing:1} to estimate local densities in an area captured by video cameras during the Hajj in Mina/Makkah in 1426H on January 12, 2006.\\
+%The Gaussian weight function used in (Wirz \textit{et al.} 2012; 2013) \cite{wirz:1}\cite{wirz:2} is introduced in Helbing \textit{et al.} (2007) \cite{helbing:1} to estimate local densities in an area captured by video cameras during the Hajj in Mina/Makkah in 1426H on January 12, 2006.\\
 
@@
 
-Weppner and Lukowicz (2013) \cite{weppner:1} estimate crowd densities during a soccer European championship public viewing event and during the Oktoberfest, by distributing volunteers in the crowd, who are carrying smart phones scanning for Bluetooth devices.
-They then use statistics to combine the different measurements in space and time.
+%In Weppner \textit{et al.} (2014) \cite{weppner:2} similar methods are applied to a city-wide festival in Zurich 2013, now supported by GPS data.
 
-In Weppner \textit{et al.} (2014) \cite{weppner:2} similar methods are applied to a city-wide festival in Zurich 2013, now supported by GPS data.
-
-*************n.b.
-in related work section put all that use wi-fi tracking, for those that use GPS tracking only mention that this does not work in indoor spaces and that it requires participation.
-Not necessary to put work on video tracking in related work - it is already mentioned in intro.
-Mention also bluetooth tracking, but bluetooth is not so ubiquitous and our principles apply also if bluetooth was used (right?) Finally, none of the mentioned work uses our approach of modeling the position of an individual as a probability distribution; our method is designed to attack the problem of having a dense crowd; we are not interested in precise estimations for freely moving crowds; rather, we are interested in obtaining precise estimation when the  concert crowds is dense and static due to this.
-blah blah (sonja: explain better)********
+Finally, none of the mentioned work uses our approach of modeling the position of an individual as a probability distribution; our method is designed to attack the problem of having a dense crowd; we are not interested in precise estimations for freely moving crowds; rather, we are interested in obtaining precise estimation when the  concert crowds is dense and static...
 

### 2016-08-02 PaperOnSmoothing/draft-31052016.tex
Started writing Positioning section.

@@
 
+The positions of smart phones are reconstructed using received signal strength (RSS) values and methods similar to multi-lateration. 
+The method is based on the decrease of signal strength when the distance between transmitter and receiver increases.
+Using a model for the decrease, the distance between transmitter and receiver can be reconstructed from the received signal strength.
+When we have at least three such reconstructed distances, the position of a transmitter can be uniquely determined using the method of circular lateration [ref: Kushki et al.].\\
+The path loss model is the Friis equation, given by
+\begin{equation}
+P_{R}=P_{T}+10\lambda\log_{10}(\frac{c}{4\pi fr})
+\end{equation}
+where we use $\lambda=2$ (line of sight condition), $c$ is the speed of light, $f$ is 2.4 GHz, and $r$ is the distance between the position $(x,y)$ (in two dimensions) of the mobile device and the position $(x_{i},y_{i})$ of router $i$, given by
+\begin{equation}
+r=\sqrt{((x-x_{i})^2+(y-y_{i})^2)}
+\end{equation}
+Received signal strengths of packets captured at different sensors are grouped in time to constitute separate `measurements'. 
+To estimate positions from these groups of measurements (signal strengths) a chi-square data fit is performed.\\
+
+The uncertainties associated with the chi-square fits are given by the covariance matrix, which is the inverse of the matrix of second-order derivatives of the chi-square function.
+The values in the covariance matrix give the uncertainty in the parameters that corresponds to a $1\sigma$ error or an increase of $\chi^2+1$.
+
 \begin{comment}
@@
 We use the existing network of Wi-Fi access points in the stadium, and capture probe signals from smart phones, which are processed and anonymised in alignment with privacy concerns.
-The positions of smart phones are reconstructed using received signal strengths and methods similar to trilateration.
+The positions of smart phones are reconstructed using received signal strength (RSS) values and methods similar to multi-lateration.
 

### 2016-08-03 PaperOnSmoothing/draft-31052016.tex
Continued writing the Positioning section.

@@
 P_{R}=P_{T}+10\lambda\log_{10}(\frac{c}{4\pi fr})
+\label{friis}
 \end{equation}
@@
 \end{equation}
-Received signal strengths of packets captured at different sensors are grouped in time to constitute separate `measurements'. 
-To estimate positions from these groups of measurements (signal strengths) a chi-square data fit is performed.\\
-
-The uncertainties associated with the chi-square fits are given by the covariance matrix, which is the inverse of the matrix of second-order derivatives of the chi-square function.
-The values in the covariance matrix give the uncertainty in the parameters that corresponds to a $1\sigma$ error or an increase of $\chi^2+1$.
+Received signal strengths of packets captured at different sensors are grouped in time to constitute separate measurements. 
+To estimate positions from these groups of measurements a chi-square data fit is performed.
+The chi-square function is minimized by setting
+\begin{equation}
+\frac{d\chi^2}{dp}=0
+\end{equation}
+The uncertainties associated with the chi-square fits are calculated using the covariance matrix of the measurements.
+If we assume Gaussian distributed measurements, the covariance matrix is the inverse of the matrix of second-order derivatives of the chi-square function, i.e.
+\begin{equation}
+C_{ij}^{-1}=\frac{1}{2}\frac{\partial^2\chi^2}{\partial p_{i}\partial p_{j}}
+\end{equation}
+The uncertainties in the parameters are obtained using a Taylor expansion of the nonlinear path loss function (Equation \ref{friis}).
+The values of the uncertainty that correspond to a $1\sigma$ error or an increase of $\chi^2+1$ are given by the projection matrix
+\begin{equation}
+P=HCH^\top
+\end{equation}
+where $H$ is a matrix consisting of (first-order) partial derivatives of the path loss model with respect to the fit parameters.
+The $1\sigma$ errors form the input to the smoothing method of the density estimation discussed in this paper.
 

### 2016-08-04 PaperOnSmoothing/draft-31052016.tex
Some changes of the Positioning section.

@@
 \end{equation}
-The uncertainties associated with the chi-square fits are calculated using the covariance matrix of the measurements.
+The uncertainties associated with the chi-square fits are calculated using the covariance matrix.
 If we assume Gaussian distributed measurements, the covariance matrix is the inverse of the matrix of second-order derivatives of the chi-square function, i.e.
@@
 The uncertainties in the parameters are obtained using a Taylor expansion of the nonlinear path loss function (Equation \ref{friis}).
-The values of the uncertainty that correspond to a $1\sigma$ error or an increase of $\chi^2+1$ are given by the projection matrix
+The uncertainty values that correspond to a $1\sigma$ error or an increase of $\chi^2+1$ are given by the covariance matrix of the parameters
 \begin{equation}
-P=HCH^\top
+C=HCH^\top
 \end{equation}
-where $H$ is a matrix consisting of (first-order) partial derivatives of the path loss model with respect to the fit parameters.
+where $H$ is a projection matrix consisting of (first-order) partial derivatives of the path loss model with respect to the fit parameters.
 The $1\sigma$ errors form the input to the smoothing method of the density estimation discussed in this paper.

### 2016-08-04 PaperOnSmoothing/draft-31052016.tex
Some changes of the Positioning section (II).

@@
 \begin{equation}
-C=HCH^\top
+C=H^\top V^{-1}H
 \end{equation}
-where $H$ is a projection matrix consisting of (first-order) partial derivatives of the path loss model with respect to the fit parameters.
+where $V$ is the covariance matrix of the measurement values, and $H$ is a projection matrix consisting of the partial derivatives of the path loss model with respect to the fit parameters.
 The $1\sigma$ errors form the input to the smoothing method of the density estimation discussed in this paper.

### 2016-08-11 PaperOnSmoothing/draft-31052016.tex
Some changes of the Positioning section (III).
Generalized the section on Positioning, nothing definite yet.

--- 2016-08-03 PaperOnSmoothing/draft-31052016.tex
@@
 
-The positions of smart phones are reconstructed using received signal strength (RSS) values and methods similar to multi-lateration. 
-The method is based on the decrease of signal strength when the distance between transmitter and receiver increases.
-Using a model for the decrease, the distance between transmitter and receiver can be reconstructed from the received signal strength.
-When we have at least three such reconstructed distances, the position of a transmitter can be uniquely determined using the method of circular lateration [ref: Kushki et al.].\\
-The path loss model is the Friis equation, given by
+The positions of smart phones are reconstructed using methods based on lateration. 
+The lateration method uses the distances between a smart phone and multiple reference points, which in this case are the existing Wi-Fi access points (APs) in the stadium.
+Smart phones transmit Wi-Fi signals which are captured at the APs.
+The captured Wi-Fi signals contain information about the mesaured signal strength.
+Using a model for the relationship between the received signal strength (RSS) of the Wi-Fi signal and the distance between transmitter and receiver, the distance between smart phone and the AP is estimated.
+The distance between the position $(x,y)$ (in two dimensions) of the mobile device and the position $(x_{i},y_{i})$ of AP $i$, given by 
+\begin{equation}
+r_{i}^2=(x-x_{i})^2+(y-y_{i})^2
+\label{distance}
+\end{equation}
+The distances define circles with radius $r_{i}$ around the APs on which the transmitter may be located.
+When we have at least three such circles, the position of a transmitter can be uniquely determined at the intersection of the circles \cite{kushki:1}.
+
+The RSS value of a signal decreases when the distance between transmitter and receiver increases.
+The decrease is given by a path loss model, given by the Friis equation
 \begin{equation}
@@
 \end{equation}
-where we use $\lambda=2$ (line of sight condition), $c$ is the speed of light, $f$ is 2.4 GHz, and $r$ is the distance between the position $(x,y)$ (in two dimensions) of the mobile device and the position $(x_{i},y_{i})$ of router $i$, given by
-\begin{equation}
-r=\sqrt{((x-x_{i})^2+(y-y_{i})^2)}
-\end{equation}
-Received signal strengths of packets captured at different sensors are grouped in time to constitute separate measurements. 
-To estimate positions from these groups of measurements a chi-square data fit is performed.
-The chi-square function is minimized by setting
-\begin{equation}
-\frac{d\chi^2}{dp}=0
-\end{equation}
-The uncertainties associated with the chi-square fits are calculated using the covariance matrix of the measurements.
-If we assume Gaussian distributed measurements, the covariance matrix is the inverse of the matrix of second-order derivatives of the chi-square function, i.e.
-\begin{equation}
-C_{ij}^{-1}=\frac{1}{2}\frac{\partial^2\chi^2}{\partial p_{i}\partial p_{j}}
-\end{equation}
-The uncertainties in the parameters are obtained using a Taylor expansion of the nonlinear path loss function (Equation \ref{friis}).
-The values of the uncertainty that correspond to a $1\sigma$ error or an increase of $\chi^2+1$ are given by the projection matrix
-\begin{equation}
-P=HCH^\top
-\end{equation}
-where $H$ is a matrix consisting of (first-order) partial derivatives of the path loss model with respect to the fit parameters.
-The $1\sigma$ errors form the input to the smoothing method of the density estimation discussed in this paper.
+where we use $\lambda=2$ (line of sight condition), $c$ is the speed of light, $f$ is 2.4 GHz, and $r$ is the distance between the mobile device and the AP.\\
+The measured RSS values contain unpredictable variation due to noise, and interference such as absorption and reflection by obstacles between transmitter and receiver (shadowing and multipath effects) \cite{goldsmith}.
+The positions of these obstacles may change in time, in particular in the case of human bodies.
+As a result, the circles in the lateration method do not intersect at a unique point, and the system of equations defined in Equation \ref{distance} do not have an exact solution.
+In this case a statistical approach is taken to the determination of the position of the smart phone.
+The least squares method is used, which in our case takes the form of a chi-square data fit \cite{bevington}.
+Based on the chi-square fit, uncertainty values for the parameter estimates are calculated.
+These values represent parameter uncertainties that correspond to a $1\sigma$ error or an increase of $\chi^2+1$. 
+The $1\sigma$ error values form the input to the smoothing method of the density estimation discussed in this paper.
 

### 2016-08-17 PaperOnSmoothing/draft-31052016.tex
Restyle of section on Brownian diffusion, to be continued.

@@
 
+\begin{comment}
 Formally, for the epochs during which a mobile device is not detected we assume Brownian motion (without drift), which is defined by
@@
 \begin{equation}
-\sigma^2=\frac{(\Delta B)^2}{\Delta t}
+\sigma^2=\frac{L^2}{\Delta t}
 \end{equation}
@@
 \marginpar{add rescaling to include information that all visitors are still inside the stadium }
+\end{comment}
+
+Formally, in case we have no new observations, we smooth the probability distribution of the previous epoch by convolution with a Gaussian probability function.
+The width $\sigma$ of the Gaussian PDF is determined through the conceptual equivalence of this approach with assuming Brownian motion of pedestrians.
+Brownian motion (without drift), is given by
+\begin{equation}
+\text{d}B(t)=\sqrt{2D}\text{d}W(t)
+\end{equation}
+where $W(t)$ is a Wiener process, and $\sigma$ is related to the diffusion constant $D$ using the relationship
+\begin{equation}
+\sigma=\sqrt{2Dt}
+\end{equation}
+The variance $\sigma^2$ is thus determined by the length $t$ of the time interval (epoch), and the diffusion constant $D$, which in turn can be determined by the walking speed of pedestrians \cite{ebeling}.\\
 

### 2016-11-15 PaperOnSmoothing/draft-31052016.tex
paper updated (positioning section)

@@
 
-This section will be written later, maybe with input from Jan.
+\marginpar{to be polished}
+We estimate the positions of smart phones using a method based on lateration. 
+The lateration method uses the distances between a smart phone and multiple reference points, which in this case are the existing Wi-Fi access points (AP's) in the stadium (see Fig.
+\ref{fig:trilateration}\footnote{Image courtesy of http://rvmiller.com/2013/05/part-1-wifi-based-trilateration-on-android/ }).
+Smart phones transmit Wi-Fi signals which are captured at the AP's.
+The captured Wi-Fi signals contain information about the measured signal strength.
+Using a physical model for the relationship between the received signal strength (RSS) of the Wi-Fi signal and the distance between a transmitter and receiver, the distance between the smart phone and the AP is estimated.
+When we have the distances from a smart phone to at least three AP's, the position of the smart phone can be uniquely determined at the intersection of three circles (Fig.
+\ref{fig:trilateration}a.) \cite{kushki:1}.
 
-The positions of smart phones are reconstructed using methods based on lateration. 
-The lateration method uses the distances between a smart phone and multiple reference points, which in this case are the existing Wi-Fi access points (APs) in the stadium.
-Smart phones transmit Wi-Fi signals which are captured at the APs.
-The captured Wi-Fi signals contain information about the mesaured signal strength.
-Using a model for the relationship between the received signal strength (RSS) of the Wi-Fi signal and the distance between transmitter and receiver, the distance between smart phone and the AP is estimated.
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=80mm]{trilateration.png}
+	\caption{Estimating position with trilateration: a) precise b) rough }
+	\label{fig:trilateration}
+\end{figure}
+\begin{comment}
 The distance between the position $(x,y)$ (in two dimensions) of the mobile device and the position $(x_{i},y_{i})$ of AP $i$, given by 
@@
 The distances define circles with radius $r_{i}$ around the APs on which the transmitter may be located.
-When we have at least three such circles, the position of a transmitter can be uniquely determined at the intersection of the circles \cite{kushki:1}.
+\end{comment}
 
@@
 The positions of these obstacles may change in time, in particular in the case of human bodies.
-As a result, the circles in the lateration method do not intersect at a unique point, and the system of equations defined in Equation \ref{distance} do not have an exact solution.
-In this case a statistical approach is taken to the determination of the position of the smart phone.
-The least squares method is used, which in our case takes the form of a chi-square data fit \cite{bevington}.
+As a result, the circles in the lateration method do not intersect at a unique point (see e.g. Fig.\ref{fig:trilateration}). 
+In this case an optimization procedure is undertaken for positioning of the smart phone.
+We use the least squares optimization method, which in our case takes the form of a chi-square data fit.
+Exact description of the method is beyond the scope of this paper and we refer the reader to \cite{bevington} for details.
+We note, however, that the statistical estimation of the position provides as also with standard deviation $\sigma$ of the estimation, which we will use in Sec.
+\ref{sec:density}.
+\begin{comment}
 Based on the chi-square fit, uncertainty values for the parameter estimates are calculated.
@@
 The $1\sigma$ error values form the input to the smoothing method of the density estimation discussed in this paper.
+\end{comment}
 
@@
 However, we also wish to preserve the uncertainty that corresponds to each estimated position.
-Therefore, we first "smooth" each position into a bivariate Gaussian distribution (`bump'), using the uncertainty values ($\sigma_{x}$ and $\sigma_{y}$) provided by the data as standard deviations.
+Therefore, we first "smooth" each position into a bivariate Gaussian distribution (`bump'), using the uncertainty values ($\sigma_{x}$ and $\sigma_{y}$) provided by the position estimation method as standard deviations.
 Then, for  $m$ we construct a two-dimensional probability density function (p.d.f.) by adding up these bumps, and  dividing the sum by $N$.\marginpar{make viz.

### 2017-01-02 PaperOnSmoothing/draft-31052016.tex
paper updated + smoothing code from philip added

@@
 \usepackage{subcaption}
+
+\newtheorem{theorem}{Theorem}
+\newtheorem{proposition}{Proposition}[theorem]
 \begin{document}
@@
 \title{Towards detecting critical crowd density with Wi-Fi positioning: Case Amsterdam ArenA}
-\author{Sonja Georgievska, Philip Rutten, Jan Amoraal, Michael Lees, Sander Klous\footnote{authors and order could change by the time the paper is submitted}}
+\author{Sonja Georgievska, Philip Rutten,  Sander Klous\footnote{authors and order could change by the time the paper is submitted}}
 \maketitle
@@
 \begin{comment} rather than estimating the position of a visitor, we create a continuous probability distribution over all possible positions.
-We rely on the law of the large numbers, with adaptation to correlated variables, to estimate the crowd density by aggregating the individual distributions.
+We rely on the law of the large numbers, with adaptation to correlated and not necessarily identically distributed variables, to estimate the crowd density by aggregating the individual distributions.
 The individual probability distribution turns gradually into a uniform distribution, with which we address the problem of outdated signals.
-Finally, we use the law of large numbers again and the "randomized address" tag to account for the fact that a portion of the MAC addresses are randomized.
+Finally, we use a weak law of large numbers again and the "randomized address" tag to account for the fact that a portion of the MAC addresses are randomized.
 Thus, \end{comment}
@@
 We borrow ideas from statistical mechanics to estimate the crowd density: rather than estimating the most likely position of a visitor in real time, we create an evolving continuous probability distribution over all possible positions (which turns out to be usually bimodal).
-We rely on the law of the large numbers, with adaptation to correlated variables, to estimate the crowd density by aggregating the individual distributions.
+We rely on the fact that we have a lot of data, to estimate the crowd density by aggregating the individual distributions.
 The individual probability distribution turns gradually into a uniform distribution, with which we address the problem of outdated signals.
-Finally, we use the law of large numbers again and the "randomized MAC address" tag, that is available with the data, to account for the fact that a portion of the MAC addresses are randomized.
+Finally, we use the fact that we have a lot of data again and the "randomized MAC address" tag, that is available with the data, to account for the fact that a portion of the MAC addresses are randomized.
 Thus, the advantage of our methodology when compared to other methodologies is that our estimation becomes more (rather than less) precise when the number of people increases.
@@
 While this does not provide us with  very useful information about the location of the MAC device, if we apply the same reasoning for all MAC devices, and we add together the spatial probability density functions for all MAC devices, we end up with a spatial distribution of the crowd density.
-If we assume that the locations of the MAC devices are mutually independent, we can  apply directly the law of the large numbers and conclude that for a dense crowd the error of the estimation of the density per square meter will vanish.
-However, we cannot assume the mentioned independence, because people tend to go to concerts in groups.
+If we assume that the locations of the MAC devices are mutually independent and identically distributed, we can apply directly the law of the large numbers and conclude that for a dense crowd the error of the estimation of the density per square meter will vanish.
+However, we cannot assume the mentioned criteria,  because people tend to go to concerts in groups.
 In this case, the variance of the estimation in the limiting case is equal to the average correlation between the locations.
@@
 We postpone the derivation of $Prob(mac_i \in R)$ a bit.
-Instead, we first show that the variance of $X/n$, that is, the variance of the proportion of devices detected in $R$ out of all detected devices, diminishes when $n$ becomes large (note that the variance of $X$ in the limiting case is out of our interest because then $E(X)$ is potentially infinite).
+Instead, we first show that the variance of $X/n$, that is, the variance of the proportion of devices detected in $R$ out of all detected devices, diminishes when $n$ becomes large (note that the variance of $X$ in the limiting case is out of our interest because in this case $E(X)$ is also potentially infinite).
 This suffices to show that our method of estimation of crowd density is theoretically sound, given the probabilities $Prob(mac_i \in R)$.
@@
 
-We can assume that there is an upper limit $\lambda$ of a number of people that go to a concert together (a group), that is, whose locations are correlated\footnote{Actually, when the crowd is so dense that people can not move freely anymore, the whole crowd becomes a "group" (as in the  Love Parade disaster) and all locations are correlated.
+It is reasonable to assume that there is an upper limit $\lambda$ of a number of people that go to a concert together (a group), that is, whose locations are correlated\footnote{Actually, when the crowd is so dense that people can not move freely anymore, the whole crowd becomes a "group" (as in the  Love Parade disaster) and all locations are correlated.
 But we aim to detect a high density with our method {\it before} this happens, to react preventively; otherwise, it is too late.}.
-Denote by $\kappa$ the maximal covariance between any $X_i$ and $X_j$ and let us write $i \sim j$ if and only if $i$ and $j$ are in the same group.
+Note that because the random variables $\{X_i\}_{i=1}^n$ take values in $\{0,1\}$, the covariances $Cov(X_i,X_j)$ take values in $[-1,1]$.
+Thus,  the covariances are right-bounded (by $1$).
+Denote by $\kappa$ the maximal covariance between any $X_i$ and $X_j$ and let us write $i \sim j$ if and only if the owners of the MAC devices $mac_i$ and $mac_j$ are in the same group of friends.
 Then, 
@@
 Let $$\mu = \frac{1}{n}\sum_{i=1}^{n} Var(X_i),$$ that is, 
-denote by $\mu$ the average variance of ${X_1, X_2, ..., X_n}$.  From \eqref{eq_Var} and \eqref{eq_covar} we have
+denote by $\mu$ the average variance of ${X_1, X_2, ..., X_n}$ (note that $\mu < 1$ from the definition of $\{X_i\}_{i=1}^n$ ).  From \eqref{eq_Var} and \eqref{eq_covar} we have
 
@@
 Note that we have greatly overestimated the covariance with the inequation in \eqref{eq_covar}, which means that in practice the variance converges to $0$ much faster than as presented.
+
+With the previous, we have actually proven a version of the law of the large numbers that can be used on other occasions. We generalize our results in the following proposition:
+
+\begin{proposition} Let $\{X_1, X_2, ...
+X_n\}$ be random variables, not necessarily independent nor identically distributed, such that they always take values in a bounded real interval.
+Suppose that the set  $\{X_1, X_2, ...
+X_n\}$ can be partitioned into subsets of maximal size $\lambda$  (a fixed constant independent of $n$), such that if $X_i$ and $X_j$ belong to different  subsets, then  $Cov(X_i,X_j) = 0$ .
+Let  $S_n = \frac{1}{n} \sum_{i=1}^{n} X_i$ Then
+
+$$ \lim_{n\to\infty} {Var(S_n)} = 0.$$ 
+\end{proposition}
 

### 2017-01-23 PaperOnSmoothing/draft-31052016.tex
changes in paper

@@
 Exact description of the method is beyond the scope of this paper and we refer the reader to \cite{bevington} for details.
-We note, however, that the statistical estimation of the position provides as also with standard deviation $\sigma$ of the estimation, which we will use in Sec.
+We note, however, that the statistical estimation of the position provides us also with standard deviation $\sigma$ of the estimation, which we will use in Sec.
 \ref{sec:density}.
@@
 In this case the packet rate is quite volatile, ranging from a few seconds to a few minutes (citation?)\marginpar{add citation}.
-This means that when we make a snapshot of the MAC addresses visible in a certain moment, we are detecting a only a very small fraction of the MAC devices.
+This means that when we make a snapshot of the MAC addresses visible in a certain moment, we are detecting only a very small fraction of the MAC devices.
 
@@
 Due to ever increasing privacy concerns and possibly other business reasons, starting from 2014, the Apple I-phones have introduced randomization of the MAC address when the device is in a probing mode (not connected to internet) (citation?)\marginpar{add citation}.
-This means that when the device switches from a probing to an online mode or vice versa, the MAC address suddenly changes and thus the device cannot be followed anymore.
+This means that devices not connected to the internet are continuously changing their MAC address and canot be followed over time.
 It is worth noting, however, that from the MAC address itself it can be determined whether the address is an original MAC address (non-randomized) or randomized.
@@
 
-In order to deal with issue 1 from above, let us start with the following observation: we are not interested in the individual locations of the MAC devices, but rather in the density of the crowd.
+In order to deal with the bimodal distribution issue  from above,  let us start with the following observation: we are not interested in the individual locations of the MAC devices, but rather in the density of the crowd.
 Also, we would  especially like to have more precise estimation when the density reaches a certain threshold, so that the crowd outside the dense region can be promptly navigated away from it.
@@
 If we assume that the locations of the MAC devices are mutually independent and identically distributed, we can apply directly the law of the large numbers and conclude that for a dense crowd the error of the estimation of the density per square meter will vanish.
-However, we cannot assume the mentioned criteria,  because people tend to go to concerts in groups.
-In this case, the variance of the estimation in the limiting case is equal to the average correlation between the locations.
-Nevertheless, we can safely assume that for a regular crowd at a concert the group size is relatively small compared to the whole crowd and thus the average correlation tends to zero as the number of people increases.
+However, we cannot assume the mentioned criteria because people tend to go to concerts in groups, that is, their locations are correlated.
+Thus, the variance of the estimation in the limiting case is equal to the average correlation between the locations.
+We will demonstrate that the average correlation still tends to zero as the number of people increases,  because for a regular crowd at a concert, the group size is relatively small compared to the whole crowd.
 Thus, the error of the estimation of the crowd density will in anyway diminish as the number of people grows.
 
-In what follows we formalize the discussion above. 
+We proceed with formalizing the discussion above. 
 Let $\{mac_1, mac_2,...mac_n\}$ be all MAC devices detected at time $t$. Let $R$ be an arbitrary region from the stadium.  Let $X_i$ be a random variable defined by
@@
 It is reasonable to assume that there is an upper limit $\lambda$ of a number of people that go to a concert together (a group), that is, whose locations are correlated\footnote{Actually, when the crowd is so dense that people can not move freely anymore, the whole crowd becomes a "group" (as in the  Love Parade disaster) and all locations are correlated.
-But we aim to detect a high density with our method {\it before} this happens, to react preventively; otherwise, it is too late.}.
+But we aim to detect a high density with our method {\it before} this happens, to react preventively; otherwise, it is too late.
+However, it is important as future work to study at which density groups start to merge, in light of studies such as \cite{density}.}.
 Note that because the random variables $\{X_i\}_{i=1}^n$ take values in $\{0,1\}$, the covariances $Cov(X_i,X_j)$ take values in $[-1,1]$.
@@
 Let $$\mu = \frac{1}{n}\sum_{i=1}^{n} Var(X_i),$$ that is, 
-denote by $\mu$ the average variance of ${X_1, X_2, ..., X_n}$ (note that $\mu < 1$ from the definition of $\{X_i\}_{i=1}^n$ ).  From \eqref{eq_Var} and \eqref{eq_covar} we have
+denote by $\mu$ the average variance of ${X_1, X_2, ..., X_n}$ (note that $\mu \leq 1$ from the definition of $\{X_i\}_{i=1}^n$ ).  From \eqref{eq_Var} and \eqref{eq_covar} we have
 
@@
 which tends to $0$ when $n \rightarrow \infty$.
-Note that we have greatly overestimated the covariance with the inequation in \eqref{eq_covar}, which means that in practice the variance converges to $0$ much faster than as presented.
+Note that we have greatly overestimated the covariance with the inequality in \eqref{eq_covar}, which means that in practice the variance converges to $0$ much faster than as presented.
 
@@
 
-\subsubsection{MAC address randomization: using Big data again}
+\subsubsection{MAC address randomization}
 
@@
 Namely, we know from the structure of the MAC address whether it has been randomized or not.
-Figure \ref{fig:randomized} shows the ratio between randomized and non-randomized addresses observed per minute from midnight until 6:00 am.
-We can see that it is quite stable through time; in fact, the Pearson correlation coefficient between the time series of randomized and non-randomized addresses is 0.88.
+Figure \ref{fig:nonrandomizedadresses} shows numbers of non-randomized and randomized adresses observed per minute from midnight until around 6:00 am.
+We can see that their ratio is quite stable through time; in fact, the Pearson correlation coefficient between the time series of randomized and non-randomized addresses is 0.98.
 
@@
 	\centering
-	\includegraphics[width=130mm]{nonrandomized.jpeg}
-	\caption{Non-randomized addresses observed per minute}
+	\includegraphics[width=100mm]{rand_nonrand.png}
+	\caption{Non-randomized (up) and randomized (down) addresses observed per minute}
 	\label{fig:nonrandomizedadresses}
 \end{figure}
+
 \begin{figure}[h!]
 	\centering
-	\includegraphics[width=130mm]{randomized.jpeg}
-	\caption{Randomized addresses observed per minute}
-	\label{fig:randomizedaddresses}
-\end{figure} 
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=130mm]{RatioRandNonrand.jpeg}
-	\caption{Ratio between randomized and non-randomized addresses observed per minute}
-	\label{fig:randomized}
+	\includegraphics[width=100mm]{ratio_hist.png}
+	\caption{Distribution of ratios between randomized and non-randomized addresses observed per minute}
+	\label{fig:ratio_dist}
 \end{figure}
 
-\noindent Thus, when estimating crowd density, we ignore the MAC devices that have a randomized address and at the end we multiply the crowd density by 1.225, to account for the ignored MAC devices.
-We choose factor 1.225 to allow for a safety margin: in the peaks of the graph in Fig.
-\ref{fig:randomized} the values are 0.225, and we would rather overestimate than underestimate the crowd density.
-\marginpar{in general: calculate it automatically, with a confidence interval}
-Note that this ratio should be re-computed periodically, to account for the changing conditions at the smart phones market. 
-\marginpar{mention the absolute numbers, or give graphs}
+\noindent Thus, when estimating crowd density, we ignore the MAC devices that have a randomized address and at the end we multiply the crowd density by a factor, to account for the ignored MAC devices.
+We choose factor 1.22 to allow for a safety margin: we would rather overestimate than underestimate the crowd density.
+The ratio  0.22 is the meam plus the standard deviation of the distribution of ratios between randomized and non-randomized addresses observed per minute (Fig \ref{fig:ratio_dist}), but in general it depends on how much underestimation/overestimation is desired.
+Note also that this ratio should be re-computed periodically, to account for the changing conditions at the smart phones market.
 
@@
 
+Make an acknowledgments section (Arena, Jessy or maybe Jessy co-author?)
+
+Mention that data is not available under law restrictions
+
+Publish code and cite?
+Also data analysis code
+
+explain exactly how the anonymization and data collection works
+
 \bibliography{references}

### 2017-02-01 PaperOnSmoothing/draft-31052016.tex
chanes in paper

@@
 Figure \ref{fig:nonrandomizedadresses} shows numbers of non-randomized and randomized adresses observed per minute from midnight until around 6:00 am.
-We can see that their ratio is quite stable through time; in fact, the Pearson correlation coefficient between the time series of randomized and non-randomized addresses is 0.98.
+We can see that their ratio is quite stable through time (Fig.
+\ref{fig:ratio_dist}); in fact, the Pearson correlation coefficient between the time series of randomized and non-randomized addresses is 0.98.
 
@@
 We choose factor 1.22 to allow for a safety margin: we would rather overestimate than underestimate the crowd density.
-The ratio  0.22 is the meam plus the standard deviation of the distribution of ratios between randomized and non-randomized addresses observed per minute (Fig \ref{fig:ratio_dist}), but in general it depends on how much underestimation/overestimation is desired.
+\marginpar{maybe show percentage of total number of addresses rather than ratio to make it bounded?}The ratio  0.22 is the mean plus the standard deviation of the distribution of ratios between randomized and non-randomized addresses observed per minute (Fig \ref{fig:ratio_dist}), but in general it depends on how much underestimation/overestimation is desired.
 Note also that this ratio should be re-computed periodically, to account for the changing conditions at the smart phones market.

### 2017-02-13 PaperOnSmoothing/draft-31052016.tex
update

--- 2017-01-23 PaperOnSmoothing/draft-31052016.tex
@@
 Thus,  the covariances are right-bounded (by $1$).
-Denote by $\kappa$ the maximal covariance between any $X_i$ and $X_j$ and let us write $i \sim j$ if and only if the owners of the MAC devices $mac_i$ and $mac_j$ are in the same group of friends.
+Denote by $\kappa \leq 1$ the maximal covariance between any $X_i$ and $X_j$ and let us write $i \sim j$ if and only if the owners of the MAC devices $mac_i$ and $mac_j$ are in the same group of friends.
 Then, 
@@
 In fact, for every MAC device that was ever observed, until it has been observed again we maintain the old probability distribution, too.
-Over time this probability distribution turns gradually into a uniform by smoothing it every epoch with a Gaussian kernel (see Fig.
+Over time this probability distribution becomes gradually uniform by smoothing it every epoch with a Gaussian kernel (see Fig.
 \ref{fig:uniform} for an example).
-The size of the kernel changes over time and is related to the maximal speed that the MAC device can have under the current density distribution of the crowd.
+The size of the kernel changes over time and is related to the maximal speed that a person can have under the current density distribution of the crowd.
 In other words, we assume a Brownian motion of the MAC device, where the speed is limited by the maximal speed of a pedestrian under the current crowd conditions (cite paper Johansson et al).
-When the MAC device is observed again, its old probability distribution is simply overwritten by the probability distribution that is computed as above.\marginpar{ref.
-picture}
+When the MAC device is observed again, its old probability distribution is simply overwritten by the probability distribution that is computed as above.
+\marginpar{most of the time the speed is close to 0.
+especially for concert crowds; it cannot be however bigger than the max speed by Johansson}
 
@@
 Namely, we know from the structure of the MAC address whether it has been randomized or not.
-Figure \ref{fig:nonrandomizedadresses} shows numbers of non-randomized and randomized adresses observed per minute from midnight until around 6:00 am.
-We can see that their ratio is quite stable through time; in fact, the Pearson correlation coefficient between the time series of randomized and non-randomized addresses is 0.98.
+Figure \ref{fig:nonrandomizedadresses} shows numbers of non-randomized and randomized addresses observed per minute from midnight until around 6:00 am.
+We can see that their ratio is quite stable through time (Fig.
+\ref{fig:ratio_dist}); in fact, the Pearson correlation coefficient between the time series of randomized and non-randomized addresses is 0.88.
 
@@
 	\centering
-	\includegraphics[width=100mm]{ratio_hist.png}
-	\caption{Distribution of ratios between randomized and non-randomized addresses observed per minute}
-	\label{fig:ratio_dist}
+	\includegraphics[width=100mm]{regRand.png}
+	\caption{Linear regression plot of randomized and non-randomized addresses observed per minute}
+	\label{fig:regRand}
 \end{figure}
@@
 \noindent Thus, when estimating crowd density, we ignore the MAC devices that have a randomized address and at the end we multiply the crowd density by a factor, to account for the ignored MAC devices.
-We choose factor 1.22 to allow for a safety margin: we would rather overestimate than underestimate the crowd density.
-The ratio  0.22 is the meam plus the standard deviation of the distribution of ratios between randomized and non-randomized addresses observed per minute (Fig \ref{fig:ratio_dist}), but in general it depends on how much underestimation/overestimation is desired.
-Note also that this ratio should be re-computed periodically, to account for the changing conditions at the smart phones market.
+This factor is derived from the slope of the linear regression fit of the two time series, which in our case turns out to be 0.2, with a standard error of 0.006.
+
+Note also that this proportion should be re-computed periodically, to account for the changing conditions at the smart phones market, but also for the fact that an increasing number of stadiums across the world offer "smart" services that motivate visitors to connect to the Wi-Fi network, which makes their MAC addresses non-randomized (fixed).
 
@@
 We need independent data sets (videos) for calibration and testing, to avoid overfitting.
-Also the paremeters should be as much as possible initially optimized from the Wi-Fi data itself and from theory.
+Also the paremeters should be as much as possible initially optimized from the Wi-Fi data itself and from theory.\marginpar{don't forget to add 'before' and 'after' videos or images}
 
